Aggressive trait selection influences life-history trade-offs, favoring survival over reproductive success in Drosophila melanogaster males

Version 1

Abstract

Life history traits are all essential traits linked to development, survival and reproduction, directly influencing individual fitness. Limited environmental or physiological resources force organisms to balance competing demands, leading to fundamental trade-offs, particularly between survival and reproduction. Male aggression often enhances reproductive success by improving resource acquisition and mate competition. However, the broader impact of high levels of aggression on individual overall fitness and life-history trade-offs remains poorly understood. To address this, we used the Drosophila melanogaster Bully line, derived from wild-type flies (Canton-S line) through artificial selection for high levels of male aggression, and assess whether increased aggressiveness influences male life-history traits. Our results reveal that increased male aggressiveness shifts the balance of a major life-history trade-off, favoring survival over reproductive success. Bully males exhibited lower mating success, shorter mating duration and less effective chemical mate-guarding. However, this was traded-off with increased lifespan, enabling more reproductive opportunities. We also uncover potential mechanistic underpinnings of the reproductive deficits. Bully males showed differences in their cuticular hydrocarbon (CHCs) profiles and transferred lower levels of cVA to females, a mate-guarding pheromone, thereby weakening their post-mating strategy. Overall, our findings demonstrate that selection for male aggression profoundly influence a key life-history trade-off highlighting aggression as an evolutionary force shaping adaptation of life-history traits and providing a foundation for future genetic and mechanistic studies.

Introduction

Life history traits, such as development time, fertility, survival, and reproductive success, are key phenotypic components that directly influence individual fitness (Braendle and Paaby, 2024; Roff, 1992; Stearns, 1992). These traits often face trade-offs: investing in one can compromise another (Fisher, 1930; Stearns, 1989; Williams, 1966). For instance, high reproductive effort may reduce survival due to the energetic costs of gamete production or increased mortality associated with mating behaviors or competition (Flatt, 2011; 2020). Trade-offs also emerge between pre- and post-copulatory strategies (Simmons, 2001; Warner et al., 1995). Males prioritizing pre-copulatory investment, such as aggressive competition for mates, may have limited resources for post-copulatory mechanisms, including sperm competition or mate guarding (Andersson, 1994; Brennan and Orbach, 2021). Understanding how such trade-offs are shaped by natural selection offers insight into how organisms optimize fitness in competitive social environments (Roff, 1992; Stearns, 1976).

The fruit fly Drosophila melanogaster (D. mel.), provides a powerful model for studying life-history trade-offs, with well-characterized developmental, behavioral, and reproductive traits (Flatt, 2020). These traits are shaped by both genetic and environmental factors and can evolve under ecological pressures. Faster development, earlier reproduction and high mating rate, for instance, are advantageous in changing environments but come at the cost of reduced longevity and fecundity (Partridge and Farquhar, 1981; Stearns, 2000). Experimental manipulations of environmental conditions such as temperature and food availability reveal how external factors shape trade-offs between survival and reproduction (Chippindale et al., 1996; Marshall and Sinclair, 2010).

Beyond development and fecundity, D. mel. serves as a model to study a wide range of social behaviors, including aggregation, aggression, and courtship, that are largely mediated by chemical communication (Kim et al., 2017; Kohl et al., 2015). Cuticular hydrocarbons (CHCs) and pheromones play diverse roles in social interaction: 7-Tricosene (7-T), a male-specific compound, increases female receptivity and modulates aggression (Grillet et al., 2006; Wang et al., 2011); cis-vaccenyl acetate (cVA), a male-transferred pheromone, attracts females, repels other males, and promotes aggregation (Kurtovic et al., 2007; Laturney and Billeter, 2016); and the female-specific hydrocarbon 7,11-heptacosadiene (7,11-HD) strongly elicits male courtship and supports species-specific mate recognition (Toda et al., 2012).

Among these interactions, male-male competition plays a central role in shaping reproductive strategies. Pre-copulatory competition involves resource defense, courtship displays, and aggressive interactions to secure mating opportunities. Courtship in Drosophila includes stereotyped behaviors such as orientation, wing extensions (producing courtship song), tapping, licking, and attempted copulation (Yamamoto and Koganezawa, 2013). Successful courtship displays lead to female acceptance, outcompeting other males that fail to meet the female’s preferences (Baxter et al., 2018; Hindmarsh Sten et al., 2025). Aggression intensifies this competition: more aggressive or dominant males often monopolize access to females (Dow and von Schilcher, 1975; Filice and Dukas, 2019; Gao et al., 2024; Nandy et al., 2016; Prunier and Trannoy, 2024). Aggressive behaviors range from wing threats and lunging to high-intensity physical confrontations like tussling and boxing (Chen et al., 2002). These behaviors promote territorial dominance and enhance mating success (Gao et al., 2024), although subordinate males may also achieve reproductive success through first access to females and prolonged copulation (Filice and Dukas, 2019).

Post-copulatory competition further shapes male reproductive success. After mating, males use multiple strategies to ensure paternity, including the transfer of seminal fluid proteins (SFPs) including Sex Peptide (SP) and Accessory gland proteins (Acps), which promote sperm storage, stimulate oviposition, and reduce female receptivity to further mating (Chapman et al., 2003; Fricke et al., 2009). These factors also contribute to the formation of a mating plug that retains sperm, acting as a passive mate-guarding strategy (Avila et al., 2011; Dunham and Rudolf, 2009). In parallel, chemical mate guarding via pheromone transfer (e.g. cVA and 7-T) reduce female attractiveness and their likelihood of re-mating (Billeter et al., 2009; Gillott, 2003; Laturney and Billeter, 2016; Miyamoto and Amrein, 2008; Scott, 1986; Zawistowski and Richmond, 1986). Male reproductive behavior is plastic: in the presence of rivals, males extend their mating duration (Bretman et al., 2009; Bretman et al., 2013; Kim et al., 2013), increase the number of sperm in their ejaculate (Garbaczewska et al., 2013), adapt seminal fuids content in their ejaculate(Hopkins et al., 2019), and present higher levels of aggression toward opponents to guard their mates (Baxter et al., 2015; Bretman et al., 2011; Dore et al., 2021; Mazzi et al., 2009). Together, this demonstrate that male-male aggression influences access to mates and reproductive outcomes through both pre- and post-copulatory competition.

While aggression can enhance male mating success in natural settings, the relationship between aggression and reproductive fitness is not always straightforward. In a previous study, Penn et al.( Penn et al., 2010) developed “Bully” males through 37 generations of selection for aggression originating from a Canton-S (Cs) background. These hyper-aggressive males frequently escalated to boxing, the most intense aggression phenotype (Chowdhury et al., 2017), but showed reduced reproductive success with virgin females compared to unselected control males (Penn et al., 2010). This suggests that although aggression may enhance competitive ability, it can also incur fitness costs. This raises important questions about the role of aggression in shaping life-history trade-offs, such as: to what extent does aggression influence life history traits related to reproduction, and does it affect the balance reproduction vs survival?

To address these questions, we investigated whether D. mel. males from the Bully line exhibit changes in life-history trade-offs compared to the unselected starting wild-type flies (Cs lines). We found that hyper-aggressive Bully males show reduced reproductive success and shorter mating duration across their lifespan, indicating lower pre-copulatory success. They also display less effective mate-guarding behavior, indicating a weaker post-copulatory strategy. However, this poor pre- and post-copulatory success was traded-off with an increased survival, which is accompanied by late-life reproductive success. We then provided a mechanistic basis for the reduced copulatory success of Bully males. They showed distinct CHCs profiles compared to Cs males, which may contribute to their lower mating success. Moreover, they transferred lower levels of cVA, a mate-guarding pheromone, to females, potentially decreasing efficiency of their post-mating strategy. Our findings underscore the complex interplay between aggression, reproductive behavior, survival, and chemical signaling, providing insights into the evolutionary dynamics of life-history trade-offs in competitive social contexts.

Results

Highly aggressive males exhibit reduced mating success and have shorter mating durations

To test the hypothesis that selection for male aggressive traits may confer mating advantages, we performed courtship competition assays in which Cs and Bully males competed for mating with virgin Cs females. No significant mating advantage was observed for Bully males (Figure 1A). We then assessed the reproductive capacity of Cs and Bully males in no-choice assays by pairing each male with a virgin Cs female for 30 minutes. Mating success was significantly lower in Bully males compared to what was observed with Cs males (Figure 1B), despite no differences in the percentage of Unilateral Wing Extension (UWE), courtship latency or mating latency, when they succeed mating (Figure 1C-E), suggesting intact courtship and mating motivation. These results were replicated using an independently selected Bully line (referred to as Bully B (Penn et al., 2010)), which also showed reduced mating success and increased aggression compared to Cs (Figure 1-figure supplement 1A, C-E). This indicates that two independent selections for male aggressive traits had simultaneously reduced males mating success.

Bully males exhibit reduced mating success and allocate less time to mating.

(A) in competitive courtship assays, the mating success of Cs and Bully males were scored when paired time with a single virgin Cs female. (B) In non-competitive courtship assays, the mating success of Cs and Bully males paired with a virgin Cs female were analyzed. (C) Proportion of time spent by Cs and Bully males doing Unilateral Wing Extension (UWE) toward a Cs virgin female in non-competitive courtship assays. (D-E) Latency to court (time difference between first interaction and first UWE) and to mate (time difference between first interaction and start of the mating) of Cs and Bully males paired with a virgin Cs female. (F) Mating duration of Cs and Bully males paired with a virgin Cs female or (G) a virgin Cs decapitated female. (H) Mating duration of Cs and Bully males that were previously raised in group of 10 males or in social isolation. For all graphs, stars indicate significant differences (* P < 0.05, ** P < 0.01, *** P < 0.001) and statistical details are given in Figure 1-table supplement 2.

Differences between Cs and Bully extended to mating duration (MD) where Bully males mated 25% shorter than Cs males (Figure 1F), a difference also observed in Bully B males (Figure 1-figure supplement 1B). This MD reduction persisted when using decapitated females (Figure 1G), ruling out female-driven termination and implicating male-intrinsic factors.

As males raised with other males have lengthened MD (Bretman et al., 2013; Kim et al., 2013), we compared singly and group-housed males for their MD. We found that both Cs and Bully group-housed males exhibited significantly extended mating durations compared to single-housed males (Figure 1H). However, the MD in Bully males remained shorter compared to Cs males in group-housed condition (Figure 1H). This indicates that Cs and Bully males respond equally to early social experience and hence do not display gene-by-environment interactions.

Together, these findings show that selection for male aggression reduces male mating success and mating duration, key traits of pre-copulatory reproductive performance. This supports our main hypothesis that selection for aggressive traits influences either or both pre- and post-copulatory strategies.

Hyper-aggressive males display reduced mate-guarding efficiency, without compromising female fertility

Given that Bully males have shorter MD and reduced mating success, we hypothesized they would compensate this poor pre-copulatory success with more efficient post-mating strategies. After mating, males typically enter a refractory period during which they are no longer aroused and do not court the females they have just mated with (Lynn et al., 2024). We first tested the hypothesis that Bully males would have a shorter refractory period than Cs males, allowing them to potentially mate with more females. To investigate this, we paired Cs and Bully males with virgin Cs females, allowed them to mate, and measured the percentage of UWE directed by these males at the female they just finished mating with over a 5-minutes post-mating period. While Cs males performed minimal UWE towards just-mated females (Figure 2A’), Bully males displayed a significantly higher percentage of UWE in this context compared to Cs males (Figure 2A’), suggesting that Bully males have a shorter refractory period than Cs males to court mated females. This effect persisted when males were swapped after mating, with Bully males showing increased UWE towards Cs-mated females (Figure 2B–B’). This indicates that Bully males present a lower refractory period to court any mated females, allowing them to potentially have higher remating rate compared to Cs males. To test this, we conducted behavioral assays in which a single Cs or Bully male was consecutively paired with 6 novel virgin females over a 3h period and scored the number of mating. Contrary to our hypothesis, Bully males did not show higher remating rates, despite consistently shorter MD (Figures 2C–D). This suggests that, despite shorter refractory period and reduced MD, Bully males did not achieve a higher remating success rate. As mating accompanies ejaculate transfer and incurs a reduction in sperm and seminal fluid reserve in D. mel. males (Hihara, 1981; Sirot et al., 2009), we hypothesized that Bully males might retain more sperm, potentially enabling them to fertilize more females sequentially within a short period. We thus analyzed female fertility after each consecutive mating and found no significant difference between Bully and Cs males (Figure 2E). Females that mated with Bully males remained as fertile as those mated with Cs males (Figure 2E). Together, our results indicate that while Bully males present shorter refractory period after mating, but their remating rate and ability to fertilize females remain comparable to those of Cs males.

Hyper-aggressive males exhibit lower mate-guarding efficiency, yet this does not affect female fertility.

(A-G). Mating duration of Cs and Bully males paired with a virgin Cs female. (A’) After mating ended, the proportion of time spent by each male to court performing UWE towards the just mated female was scored for 10 minutes. (B’) The proportion of time spent by the males performing UWE an unfamiliar mated female was scored for 10 minutes. (C) Total number of mating reached by Cs and Bully males over six consecutive mating with virgin Cs females. (D) Mating duration of Cs and Bully males over six consecutive mating with virgin Cs females. A significant interaction between males’ genotype and number of mating was found, indicating that Cs decreased their mating duration with the number of mating, while Bully males did not. (E) Proportion of mating that gave rise to progeny in consecutive mating assay. (F’) Proportion of time spent by naïve Cs and Bully males to display UWE either toward Cs-mated or Bully-mated decapitated female over an observation period of 2 minutes. (G’) Proportion of Cs-mated and Bully-mated females that re-mated 12 or 14 days after a first mating. For graphs A, A’, B, B’, F, F’, G and G’, stars indicate significant differences (* P < 0.05, ** P < 0.01, *** P < 0.001) and statistical details are given in Figure 1-table supplement 2.

Due to a shorter MD in Bully males, we reasoned that a lower quantity of CHCs, acting as mate-guarding pheromones, may be transferred to the females’ body, making Cs females that mated with Bully males more attractive to other males compared to those mated with Cs males. To assess female attractiveness independently of any confounding effects of female receptivity, we performed preference assays for decapitated females, who lack receptivity. Cs females were first mated with either Cs or Bully males, then decapitated and simultaneously presented to naïve males in a binary choice assay (Figure 2F). Both Cs and Bully males preferentially courted decapitated females that had previously mated with Bully males (Figure 2F’, green side), showing that females mated with Bully males remain more attractive than those mated with Cs males. In addition to mate guarding pheromones, males transfer seminal fluids that affect long term female receptivity, which may differ between Bully and Cs males. We therefore conducted mating assays in which virgin Cs females were initially mated with either Cs or Bully males. Several days later, these mated females were paired with naïve virgin Cs males, and the proportion of females that remated was analyzed (Figure 2G’). Females previously mated with Cs males, remated significantly less than females that had mated with Bully males (Figure 2G’, green), further supporting weaker mate-guarding by Bully males.

We conclude that Bully males fail to offset their poor pre-copulatory success with more efficient post-mating strategies. Their mates present a higher remating rate, likely due to reduced transfer of mate-guarding pheromones and seminal fluid. Thus, selection for aggressive traits compromises both pre- and post-copulatory male performance.

Hyper-aggressive males present differences in their CHCs profiles and transfer lower quantities of mate-guarding pheromones to females

To identify mechanisms underlying reduced mating success (Figure 1) and mate-guarding (Figure 2F-G) in Bully males, we reasoned that these could be attributable to differences in males and females CHCs profiles. To test this, we quantified the CHCs profiles of naïve Cs and Bully males. PCA analysis revealed distinct CHCs profiles between naïve Cs and Bully males (Figure 3A). Comparing individual compounds of naïve Bully males relatively to Cs ones, we found that Bully males exhibited significantly higher levels of alkanes, monoenes, and methyl-alkanes (Figure 3B, Table 1). These differences may contribute to modulating female attractiveness toward males.

Hyper-aggressive males exhibit altered CHCs profiles and transfer reduced amounts of mate-guarding pheromones to females.

(A) PCA analysis on mean concentration for each Cuticular Hydrocarbon (CHCs) compound of naïve Cs and Bully. Their CHCs profiles were found similar. (B) Relative concentrations of Bully’s CHCs were calculated with respect to those of Cs males. The CHCs levels in Cs males were normalized to 1, and Bully’s CHC concentrations were expressed relative to this reference. (C) PCA analysis on mean concentration for each CHCs compound of virgin, Cs- and Bully-mated females. (D) Relative concentrations of CHCs compounds in mated females compared to virgin females. CHCs levels in virgin females were normalized to 1, and the concentrations in mated females were expressed relative to this baseline. Significant differences are indicated by letters: a indicates a significant difference between virgin and Cs-mated females; b indicates a significant difference between virgin and Bully-mated females; and c indicates a significant difference between Cs- and Bully-mated females (a P < 0.05; aa P < 0.01; aaa P < 0.001). (E) Mean ± SE concentration in ng of CHCs compounds that were not present in virgin females. (F) Mating duration of Cs and Bully males with a Cs female. (F’) cVA concentration measured within Cs- and Bully-mated genitalia tract right after mating. For graphs E, F and F’, stars indicate significant differences (* P < 0.05, ** P < 0.01, *** P < 0.001) and statistical details are given in Figure 1-table supplement 2.

Our results suggest that the higher remating rate in females mated with Bully males (Figure 2G’) may be due to increased attractiveness compared to females mated with Cs males (Figure 2F’), a process influenced by pheromones transferred from males to females such as the blend of cVA/7-T (Laturney and Billeter, 2016). To test this hypothesis, we extracted CHCs from the cuticles of Cs virgin and Cs females mated to either a Cs or Bully males and compared their CHCs profiles. As a first step, PCA analysis showed that Cs virgin females exhibited distinct CHCs profiles from those of both Cs and Bully-mated females (Figure 3C). The first (PC1) and the second (PC2) dimensions explain 80.7% and 11.1% of the variance in CHCs profiles, respectively. High cos² values indicate a strong correlation between CHCs compounds and the principal components, suggesting that the PCA captures the major part of the variance in the data (Figure 3-figure supplement 1A). In addition, some compounds such as 2MeC24, C23:1(6), cVA and C25:1(6) strongly influence PC1 and explain the difference between the clusters, i.e. female mated status (Figure 3-figure supplement 1B-D). Individual compound analysis revealed that this difference was primarily due to the transfer of male-specific CHCs during mating not found in virgin females (Figure 3D and E, Table 2). Relatively to virgin females, C23:1(9) (9-Tricosene, 9-T), C23:1(7) (7-Tricosene, 7-T), C23:1(6), C25:1(6) and cVA were significantly elevated in mated females (Figure 3D and E, Table 2). These results are consistent with previous studies showing that 7-T and 9-T are acquired by females during mating (Everaerts et al., 2010; Laturney and Billeter, 2016). When comparing the two groups of mated females, individual compound analysis showed that C23:1(5) (5-Tricosene, 5-T) and C25:1(5) (5-Pentacosene, 5-P) were significantly elevated in Bully-mated females compared to Cs-mated females (Table 2, surrounded in red, Figure 3D). These findings suggest that 5-P and 5-T might contribute to promoting remating in females that have previously mated with Bully males (Figure 2F’ and G’). Given that Bully males also showed higher levels of both 5-P and 5-T compared to naïve Cs males (Figure 3B), it is likely that the elevated levels of these compounds observed in females result from their transfer during mating.

cVA, a male-specific compound that acts as an anti-aphrodisiac and a mate-guarding molecule, is primarily transferred to the female’s genital tract during mating, with only trace amounts deposited on the cuticle (Laturney and Billeter, 2016). Females with higher cVA levels (Kurtovic et al., 2007), combined with 7-T, are less attractive to males (Laturney and Billeter, 2016). Given that remating rate of Bully-mated females was higher than that of Cs-mated females (Figure 2G’), we hypothesized that Bully-mated females would receive less cVA in their genital tract during mating, thus decreasing their attractiveness than when mating with Cs males. To test this, we conducted mating assays with Cs females paired with either Cs or Bully males, followed by dissection of the females’ genital tracts to quantify the amount of cVA transferred. We found that females mated with Bully males had about three times lower cVA quantities in their genital tracts compared to those mated with Cs males (Figure 3F’). Combined with higher quantities of 5-T and 5-P, reduced cVA level in genital tract likely contribute to the increased attractiveness and higher remating rates observed in females that previously mated with Bully males.

Overall, our findings suggest a mechanistic basis for the reduced pre- and post-copulatory behaviors observed in Bully males. Differences in CHCs profiles between naïve Cs and Bully may underlie a reduced attractiveness to females, thereby reducing their mating success. Additionally, increased levels of 5-T and 5-P, combined with reduced cVA transfer to the female genital tract, may contribute to the lower mate-guarding efficiency observed in these hyper-aggressive males.

The lower copulatory strategy of aggressive males is offset by a longer lifespan and maintenance of reproductive success later in life

Based on the main life-history trade-off between reproduction and survival, we hypothesized that the reduced pre- and post-copulatory success in Bully males would increase their survival. To test this, we conducted longitudinal courtship assays in which individual Cs and Bully males were paired weekly with a new virgin Cs female until their natural death. Throughout their lifespan, Bully males consistently displayed shorter MD than Cs males (Figure 4A). While Cs males increased their MD with age or experience, a pattern reported in another Drosophila species (Dhole and Pfennig, 2014), this effect was absent in Bully males, whose MD remained stable across mating and throughout their lives (Figure 4A). Analysis of individual male slopes confirmed this divergence: most Cs males showed increasing MD across mating events, while Bully males did not (Figure 4—figure supplement 1). Mating success followed a similar pattern. Although Bully males mated less frequently than Cs males during the first 70 days, they maintained reproductive activity after all Cs males had ceased reproductive activity. All Cs males stopped mating by day 78 and died by day 97, whereas Bully males continued to mate until day 98 and survived up to 119 days (Figure 4B). Remarkably, these late-life matings produced viable offspring, demonstrating that Bully males retain fertility at older ages.

Aggressive males offset their reduced reproductive success with a longer lifespan and preserved reproductive success in later life.

(A) Mating duration of Cs and Bully males with virgin Cs females across their lifespan. A significant interaction between male genotype and age was observed: Bully males maintained a stable mating duration over time, whereas Cs males showed an increase in mating duration with successive mating events. (B) Number of mating throughout the lifespan of Cs and Bully males. Numbers next to data points indicate the proportion of mating. Arrows denote the last mating event and the time of death for each genotype. The number of offspring produced from the final Bully mating is indicated next to the corresponding arrow. (C) Survival curves of progeny from ♀Cs × ♂Cs and ♀Cs × ♂Bully crosses. Progeny from Bully males had a median survival (EC₅₀) of 79 days, with the last individual dying at day 106 (green line), while progeny from Cs males showed an EC₅₀ of 57 days, with the final death at day 90 (yellow line). (D) Survival curves of males raised in group of 12♀Cs & 10♂Cs or 12♀Cs & 10♂Bully. Bully males had a median survival (EC₅₀) of 56 days, with the last individual dying at day 71 (green line), while Cs males showed an EC₅₀ of 43 days, with the final death at day 67 (yellow line). Statistical details are given in Figure 1-table supplement 2.

Bully males survived significantly longer than Cs males, suggesting a shift in the reproduction-survival trade-off. To explore this further, Cs virgin females were mated with either Cs and Bully males, and progeny survival was monitored. Progeny from Bully mating showed a median survival (EC50) of 79 days, with the last male dying at day 106 (Figure 4C, green), while progeny from Cs males showed an EC50 of 57 days, with the last death occurring at day 90 (Figure 4C, yellow). This showed that the median survival was extended by ∼22 days (79 vs. 57 days) in Bully males. We then monitored male survival in mixed-sex groups composed of ten males and twelve females. Bully males still showed increased longevity, with an EC50 of 56 days and the last death on day 71, compared to Cs males, that had an EC50 of 43 days and all died by day 67 (Figure 4D). These findings are consistent with our previous results (Figure 4B and C), confirming that Bully males have a longer lifespan and can transmit this trait to their offspring.

Our findings demonstrate that males selected for high levels of aggression show reduced investment in mating success but exhibit extended lifespan and sustained reproductive output later in life. Together, these findings support our initial hypothesis that selection for aggressive traits influences a key life-history trade-off, by favoring survival at the expense of reproductive traits.

Discussion

Our study reveals that selection for aggressive traits in D. mel. males influences one key life-history trade-off, in which survival is extended while reproductive success is reduced. Our results indicate that selection for aggressive traits directly or indirectly affects both pre-and post-copulatory strategies in Bully males, challenging the hypothesis that aggression necessarily enhances mating success (Andersson, 1994; West-Eberhard, 1983). Aggressive males had lower pre-copulatory success with decreased mating success and shorter mating duration than their less aggressive counterparts. Despite reduced refractory courtship period after mating, Bully males fail to achieve higher remating rate and exhibit a less effective mate-guarding strategy, as evidenced by the increased remating rate and potentially attractiveness of their previous mates. Differences between Bully and Cs males also extend to chemical profiles, which may account for reduced mating success in Bully males. In addition, females mated to highly aggressive males exhibited quantitative differences in CHCs profiles compared those mated with Cs males, along with a dramatic reduction in cVA quantities in their genital tracts after mating, providing a mechanism underlying a less efficient mate-guarding. Our findings suggest that artificial selection applied to male aggressive traits to produce the Bully line, is associated with changes in reproductive traits, enhanced survival, and increased CHCs levels, suggesting the existence of pleiotropic genes whose variation links aggression with these other traits.

The increased survival rate observed in highly aggressive males raises questions about the evolutionary consequences of aggression. While Bully males exhibit reduced pre- and post-copulatory success compared to Cs males, their extended lifespan allows them to sustain mating activity and produce viable offspring at later ages, challenging the classical view that early reproductive success is the primary driver of male fitness (Flatt, 2011). This shift suggests that alternative reproductive strategies, even those associated with lower mating success, can persist if they confer long-term survival advantages. Moreover, the extended lifespan of both Bully males and their progeny suggests the transmission of pleiotropic genetic traits that influence both aggression and survival, or the transmission of independent genetic loci affecting these traits separately. A previous study identified a limited number of genes that are differentially expressed in the Bully line compared to Cs (Chowdhury et al., 2017). One of these genes, CG13646, a putative transmembrane transporter, is down-regulated in Bully males compared to Cs. Future research should investigate the genetic and neurobiological mechanisms underlying the trade-off reproduction/survival by examining the roles of candidate genes whose expression is up- or down-regulated in the Bully line.

Despite maintaining high courtship intensity, highly aggressive males exhibit reduced mating success and mating duration, suggesting that aggression is linked to physiological or behavioral traits that may be suboptimal for reproductive performance. For example, males might be less attractive for reasons unrelated to their courtship efforts. Rather than maximizing immediate reproductive output, aggressive males might allocate energy in a way that prioritizes long-term survival, potentially reflecting a life-history shift toward extended reproductive capacity at the cost of reduced mating success.

We showed that naïve highly aggressive individuals present distinct chemical profiles compared to unselected wild-type flies, indicating that selection for aggressive traits may have broader impacts, such as indirectly influencing mate choice through male attractiveness to females. Elevated levels of most CHCs in aggressive males suggest that they could be perceived as more attractive to females than wild-type males. However, despite these chemical traits potentially enhancing female attraction, aggressive males exhibit lower mating success, highlighting that a higher quantity of CHCs alone does not guarantee mating success. However, besides 7-T as CHCs fostering male attractiveness, other cuticular compounds, such as palmitoleic acid signaling through Or47b, modulate courtship behaviors in an age-dependent manner (Dweck et al., 2015; Kohlmeier et al., 2021; Lin et al., 2016). This mechanism enables older males to initiate courtship more rapidly and with greater intensity, potentially enhancing their mating success relative to younger males (Lin et al., 2016). In our experiments, we did not measure palmitoleic acid levels in Bully and Cs males. However, it remains possible that Bully males have lower levels of palmitoleic acid compared to Cs males. This could explain their decreased mating success by potentially reducing their attractiveness. Aggressive males may signal their fitness through chemical cues, but their high level of aggression could hinder mate choice by females, who may prioritize other factors, such as male courtship behavior or compatibility in post-mating interactions, over chemical signaling alone. This raises the possibility that aggression-driven sexual selection operates through both direct behavioral competition and indirect chemical signaling mechanisms, influencing overall reproductive success.

The influence of aggression extends beyond accessing mates, affecting post-mating behaviors, mate-guarding and chemical communication, shaping male reproductive strategies and female remating dynamics. Hyper-aggressive males exhibit a shorten refractory period to court females showed by an elevated post-mating courtship yet reduced mate-guarding efficiency, likely contributing to increasing remating rate of both males and females respectively. This suggests that while aggression may provide advantages in direct competition, it can incur reproductive costs by diminishing mate-guarding efficiency. The chemical analysis further supports this assumption, showing that females mated with the hyper-aggressive Bully males exhibit elevated levels of specific CHCs (5-T and 5-P) and significantly lower cVA transfer into the female genital tract, which may fail to efficiently decrease their attractiveness, thus triggering higher remating rate. This is supported by other study showing that lower cVA levels can shorten the effectiveness of chemical mate-guarding, allowing quicker female re-attractiveness (Ejima, 2015). As 7-T and cVA act in concert to modulate behavior (Laturney and Billeter, 2016; Verschut et al., 2023), it is possible that the elevated levels of 5-T and 5-P in Bully-mated females interact synergistically too, or with potential other CHCs, to fail to reduce female attractiveness. This would suggest that aggression may indirectly influence post-copulatory sexual selection by changing female chemical signatures.

In summary, our findings emphasize the need to consider aggression within a broader life-history and evolutionary framework. The influence of aggression on the trade-off survival/reproduction, potentially through chemical signaling, illustrates how selection for competitive traits does not occur in isolation but is shaped by interrelated behavioral and physiological factors. This supports the notion that aggression benefits to individuals in competitive environment but may induce reproductive constraints. By framing aggression as part of a broader life-history strategy rather than an isolated trait, our study highlights the intricate balance between competition, survival, and reproductive success. These insights offer novel perspectives on the evolutionary consequences of aggressive behavior and its role in shaping reproductive strategies in animal populations.

Material and methods

Experimental model and subject details

Fly stocks

Drosophila melanogaster strains were raised at 25°C under a 12h:12h light/dark cycle (LD= 9:00a.m. – 9:00p.m.) on standard fly food (74% corn flour, 28% yeast, 40% sugar, 8% agar, 20% Moldex). Canton-S (Cs) flies (from Edward Kravitz’s laboratory, RRID:BDSC_64349) were used as wild-type. The Bully A and B lines were originally selected in Edward Kravitz’s laboratory from Cs during 37 generations, resulting in hyper-aggressive male flies (Garbaczewska et al., 2013; Hopkins et al., 2019).

Experimental arena

Experimental arenas were described in details in Trannoy et al., 2015 (Trannoy et al., 2015). Briefly, resin blocks containing three circular behavioral arenas (dimension: 2.3 cm diameter, 1.7 cm height) were used to conduct behavioral assays. Each arena was divided in two equal sizes by a plastic divider allowing flies to acclimate without physical interactions with the second fly. Then, dividers are removed from the arenas to start the behavioral assays. All behavioral assays were performed in empty arenas, expect for aggression assays and courtship assays with decapitated females, in which a food cup (1.5 cm diameter, 1 cm height) containing fresh standard fly food with a drop of fresh yeast paste on the surface was placed in the center of each arena.

Method details

Commonalities between all behavioral assays

Late-stage pupae were collected from stocks raised at 25°C. Male pupae were individually placed in 5ml vials (Dutscher, ref 390597) containing about 1ml of standard fly food while female pupae were placed in group of 15 in ≈47ml vials (Dutscher, ref 789001B) containing about 15ml of standard fly food. All vials were placed at 25°C for 7 days. Seven-day-old individuals were randomly inserted in arenas by negative geotaxis, and allow to acclimate for 5 min before experiments. All behavioral assays were performed at 25°C. All tests were recorded with Basler camera Ace1970 through the StreamPix8 multi video recording software (norpix.com) and analyzed with BORIS software (Friard and Gamba, 2016). After 15 min or 30 min, if pairs of flies had not yet lunged or mated respectively, we considered that no aggression or mating occurred and arenas were not considered for analysis. No power analyses were performed prior to the experiments, as we chose to use sample sizes of approximately 20 replicates, which is considered standard practice in the field.

Aggression assays

Two days prior to the aggression assays, males were anesthetized with CO₂, and half were marked with a small white dot on the thorax before being returned to their original isolation vials. Aggression tests were conducted between ZT0 and ZT3 using size-matched male pairs. Males competed for access to a central food cup containing yeast. The following behaviors were recorded for 10 minutes following the first lunge: latency to lunge (i.e., the time between the initial encounter on the food cup and the first lunge), total number of lunges, and number of boxing events.

Mating assays

Seven-day-old individuals were tested in arenas without any food source. Latency to court and to mate, defined as the time between the first encounter and either the first Unilateral Wing Extension (UWE) or the initiation of mating, respectively, were measured. Total UWE duration was recorded over a 10-minute period starting from the first encounter. Mating proportion and mating duration were also scored. In all assays, males were reared in social isolation, except for Figure 1H, where they were housed in groups of 10 males for 7 days.

Mating assays with decapitated females

Females were anesthetized with CO₂ and placed under the binocular for head removal. Experiments with decapitated females were performed in arenas with food cup filled with fresh fly food but without yeast.

Multiple consecutive mating events assays

Single naïve Cs and Bully males were sequentially paired with six virgin Cs females. For each male, mating duration and the number of successful mating were recorded. After each copulation, the mated female was transferred to a 5 ml vial containing fly food to assess progeny production.

Mating assays throughout the lifespan

24 Cs and Bully males were reared in isolation and paired with seven-day-old virgin Cs females for mating assays, conducted every seven days until all males had died. At each time point, mating duration and mating success (proportion of mating) were recorded. After all Cs males had died, females that had mated with Bully males were transferred to 5 ml vials containing fly food to assess progeny production.

Attractiveness assays

Measure of post-mating UWE

Mating assays were performed by pairing Cs and Bully males with Cs females. Right after mating end, we scored the total time that males displayed UWE for 10 minutes towards the just-mated females.

Measure of post-mating UWE with switched females

Mating assays were performed with naive Cs or Bully males and virgin Cs females. Immediately after mating ends, females were switched: Cs males interacted with Bully-mated females and Bully males interacted with Cs-mated females. Then total time of UWE towards the females was scored during 10 min.

Measure of post-mating UWE with decapitated females

Mating assays were conducted using naïve Cs or Bully males paired with virgin Cs females. Following copulation, mated females were decapitated. One decapitated Cs-mated and one Bully-mated female were then placed on the surface of the food cup. A new naïve Cs or Bully male was introduced and allowed to interact with both decapitated females. The duration of Unilateral Wing Extensions (UWE) directed toward each female was recorded over a 2-minute period following the first encounter. Arenas in which males interacted with only one of the two females were excluded from the analysis, as no choice could be determined.

Re-mating assays

Mating assays were performed with naïve Cs or Bully males and virgin Cs females. Mated-females were then individually placed in 5ml vials containing about 1ml of standard fly food and kept at 25°C. 12 or 14 days after the initial mating, mated-females were paired with a new naïve Cs male in empty arena and the proportion of re-mating during 30 min was scored.

Survival assays

Survival with isolated males

Late-stage pupae from the progeny of ♀Cs × ♂Cs and ♀Cs × ♂Bully crosses were individually isolated in 5 ml vials containing 1ml of standard fly food and kept at 25 °C until adult emergence. The number of dead individuals was recorded daily until all flies had died. To maintain food quality, flies were transferred weekly to fresh vials containing approximately 1 ml of new fly food.

Survival with group-raised males

Late-stage Cs and Bully male pupae and Cs female pupae were placed in group of 12♀Cs-10♂Cs or 12♀Cs-10♂Bully in ≈47ml vials containing 15mL standard fly food and kept at 25 °C until adult emergence. The number of dead males was recorded daily until they all had died. To maintain food quality, flies were transferred twice a week to fresh vials containing approximately 15 mL of new fly food.

CHCs analysis

Extraction and analysis of cuticular hydrocarbons (CHCs) of naive Cs and Bully males, as well as virgin Cs, Cs- and Bully-mated females were performed in Jean-Christophe Billeter’s lab, Groningen, Netherlands, with a Gas Chromatography coupled with Flame Ionization Detection (GC-FID) as already described (Laturney and Billeter, 2016). Briefly, single whole flies were put in a 2ml glass vial (Supelco certified vial kit, 2ml clear glass vial 12×32mm and Supelco inserts) containing 50µl of nC18+nC26 10 ng/µl standard solution (made in Billeter’s lab) diluted in n-Hexane 99%+ (Acrōs Organic®). Vials were then vortexed at the lowest speed for 2min, afterward flies’ body were gently removed from vials using a “J-shape” paper clip sliced at the end, taking care not to damage the body to prevent from hemolymph contamination. Finally, samples were placed in GC-FID for apolar compounds analysis only.

cVA quantification into genital tract

Mating assays were performed with naïve Cs or Bully males and virgin Cs female. Right after mating, females were transferred into a 5ml vial previously put in ice in -80°C fridge to make them fall asleep quickly and prevent from sperm ejection. Mated-females’ genital tract was dissected in PBS with precision forceps under a binocular loupe. Protocol was already employed in (Verschut et al., 2022). Briefly, abdomen was torn to make ovaries and genital tract out and visible, then ovaries, oviduct, spermathecae, parovaria and cuticula were removed to keep only the uterus, taking care not to injure it to prevent from sperm leak. Each uterus was inserted in a 2ml glass vial (Supelco certified vial kit, 2ml clear glass vial 12×32mm and Supelco inserts) containing 50µl of n-Hexane 99%+ (Acrōs Organic) for 1h at ambient temperature. Hexane solution was then evaporated under helium flow to prevent from cVA oxidation with dioxygen. Samples were sent to Jean-Christophe Billeter’s lab for GC-FID analysis of cVA, as describe above in CHCs analysis method part.

Quantification and statistical analysis

No data have been removed from the statistical analysis. Some experiments were independently replicated by different authors to reduce potential bias. All statistical analyses were performed on R software version 4.0. (Team, 2020) with packages (Fox and Weisberg, 2019), coin (Hothorn et al., 2006), dplyr (Wickham et al., 2023), factoextra (Kassambara and Mundt, 2020), FactoMineR (Lê et al., 2008), lme4 (Bates et al., 2015), lmerTest (Kuznetsova et al., 2017), MASS (Venables and Ripley, 2002), MuMIn (Bartoń, 2023), nlme (Pinheiro et al., 2022), performance (Lüdecke et al., 2021), pgirmess (Giraudoux. P, 2018), predictmeans (Luo et al., 2021), survival (Therneau, 2022), vegan (Oksanen J, 2022).

Comparisons of means (k =2) were performed with a T-test when possible (i.e. mating duration) otherwise a Fisher Pitman permutation test was used (i.e. UWE, latency to court and to mate). For comparisons with more than 2 conditions, a linear model (LM) was used and tested either with a type II ANOVA when possible (i.e. mating duration) or a permutation test instead (i.e. latency to lunge and CHC concentrations). For comparison of proportions and counting events, a type II analysis of deviance table with a Likelihood Ratio Test (i.e. mating success, proportion of female re-mating, number of lunge and boxing events) or a Wald Chi-square test (i.e. number of consecutive mating events) was performed. A Principal Component Analysis (PCA) was performed to see different CHC profiles and tested with a PerMANOVA and a SIMPER analysis. Survival data were analysed with a Weibull regression model and a Likelihood Ratio Test.

When paired data, dependency was considered thanks to linear mixed model (LMM) with the source of dependency as random factor. Data following a distribution other than a normal distribution (e.g. binomial or Poisson distribution) were computed in a generalized linear model (GLM) (i.e. mating success, number of consecutive mating events, number of mating giving progeny, proportion of females re-mating, number of lunges and boxing events). When necessary, interactions were considered within models and tested. All post-hoc comparisons were performed with a sequential Bonferroni correction on the alpha level using either Holm method (Holm, 1979) when few numbers of comparisons or Benjamini and Hochberg method (Benjamini and Hochberg, 1995) when more comparisons. All statistical details are given in Table S1.

eLife Assessment

This valuable study addresses the effects of selection on aggression on fitness and life-history trade-offs in Drosophila melanogaster. However, the evidence presented is incomplete and does not support the claims proposed in the study of increased survival of highly aggressive males at the expense of reproductive success and shorter mating duration. The main limitation of the study is the choice to use males from only one aggressive Drosophila line in combination with CantonS females, that do not allow disambiguation between nonaggression-related factors, such as hybrid vigor and aggression-related factors influencing mating and lifespan.

Reviewer #1 (Public review):

Summary:

This study asks how selection for male aggressiveness affects life-history and reproductive fitness traits in Drosophila melanogaster males.

Strengths:

Multiple comprehensive assays are used to address the question.

Weaknesses:

(1) The flies used for comparisons are inadequate. Behavioral assays compare Bully males mated to non-coevolved Cs females with Cs males mated to coevolved Cs females.

(2) Lifespan analysis is done on male progeny of Cs females mated to either genetically more distant Bully or co-evolved Cs males; the longer lifespan and performance on the former is interpreted as a trade-off with aggressiveness, rather than a simple explanation of hybrid vigor.

(3) Differences in CHCs between Bully and Cs males and Cs females mated to those males are not shown to cause differences in measured behavioral outcomes.

Reviewer #2 (Public review):

Summary:

The authors compare "Bully" lines, selected for male aggression, to Canton-S controls and find that Bully males have lower mating success, shorter mating durations, and remate sooner. Chemical analyses show Bully males have distinct cuticular hydrocarbons (CHC) signatures and transfer markedly less cVA to females, offering a plausible mechanistic link to weaker mate-guarding.

Paradoxically, Bully males live longer and remain fertile at older ages when CS males no longer mate, indicating a shift in the reproduction-survival trade-off in aggression-selected populations.

Importantly, the work sheds light on proximate mechanisms, demonstrating that shifts in CHCs and pheromone transfer co-occur with changes in fitness traits, thus offering new entry points for understanding life-history evolution.

Strengths:

The manuscript's strengths lie in its comprehensive and integrative approach framed within an evolutionary context. By combining behavioral assays, chemical profiling, and lifespan measurements, the authors reveal a coherent pattern linking aggression selection to life-history trade-offs. The direct quantification of cVA in female reproductive tracts after mating provides a particularly compelling mechanistic correlate, strengthening the link between behavior and chemical signaling. Findings on altered 5-T and 5-P levels further highlight how chemical communication shapes mating and mate-guarding strategies. Analytical approaches are largely rigorous, and the results provide valuable insights into the pleiotropic effects of selection on socially relevant traits. The study will be of interest to Drosophila biologists working on sexual selection, behavioral evolution, and aging.

Weaknesses:

The weaknesses are primarily conceptual rather than procedural. The generality of the findings is uncertain, as selection appears to be represented by only one (and a second closely related) Bully line, limiting conclusions about selection responses versus line-specific drift or founder effects. The causal link between aggression selection and increased longevity is not established: the data show a correlated shift but do not identify mechanisms underlying lifespan extension. In several places, the manuscript uses causal language (e.g., that selection 'influences' longevity or mating strategy) where association would be more accurate; this should be toned down to avoid overstatement. Ecological relevance is also not addressed, since laboratory conditions may bias the balance between costs and benefits of aggression compared with variable natural environments. Addressing these points would strengthen both the impact and clarity of the study.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment:

This valuable study addresses the effects of selection on aggression on fitness and life-history trade-offs in Drosophila melanogaster. However, the evidence presented is incomplete and does not support the claims proposed in the study of increased survival of highly aggressive males at the expense of reproductive success and shorter mating duration. The main limitation of the study is the choice to use males from only one aggressive Drosophila line in combination with Canton-S females, that do not allow disambiguation between nonaggression-related factors, such as hybrid vigor and aggression-related factors influencing mating and lifespan.

We would like to clarify the points raised in the eLife assessment.

The report states that we relied on a single line of hyper-aggressive males tested with Canton-S females, and implies that Bully and Cs have not co-evolved. This is a misunderstanding: Bully flies were derived from Cs population. Thus, Bully and Cs have co-evolved. In addition to the Bully A line presented in the main figures of the manuscript, we replicated several of our findings with a second independent selected line, Bully B. Results from courtship assays involving both Bully A and Bully B couples males and females were presented in Figure Supp1. We apologies for not having made this more explicit in the original manuscript, which we will correct. These experiments should alleviate the concerns from the reviewers; they demonstrate that our conclusions are supported by two independent hyper-aggressive lines, and these include assays with selected male and female flies.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This study asks how selection for male aggressiveness affects life-history and reproductive fitness traits in Drosophila melanogaster males.

Strengths:

Multiple comprehensive assays are used to address the question.

We thank the reviewer for recognizing these strengths.

Weaknesses:

(1) The flies used for comparisons are inadequate. Behavioral assays compare Bully males mated to non-coevolved Cs females with Cs males mated to coevolved Cs females.

We thank the reviewer for this comment, which made us realize that we had not sufficiently highlighted some of our experiments. The Bully lines used in our work were derived from Canton-S flies and thus did co-evolve with Cs. As originally described by Penn et al. (2010), highly aggressive “Bully” lines were generated through selective breeding from Canton-S males that consistently won aggressive encounters. After 34–37 generations, stable Bully lines were established. Thus, 1) Bully and Cs flies have co-evolved and 2) the selection applied was male-specific. Independent selection replicates produced distinct lines, including Bully A and Bully B. Previous studies only characterized Bully A (Penn et al., 2010; Chowdhury et al., 2017), but our work includes both Bully A and Bully B (Fig. S1).

The rationale for pairing Bully or Cs males with Cs females (with which both male types co-evolved) follows the approach used by Dierick et al. (2006), who investigated how the male-specific selection for aggression affected courtship and mating behaviors by testing them with standard Canton-S females. This design allows to isolate the effects of male genotype and behavior on courtship and mating outcomes, avoiding confounding effects from female behavioral changes.

We initially compared selected Bully pairs (Bully males × Bully females) (Fig. S1) with Cs pairs and observed similarly shortened mating durations in both Bully × Bully and Bully × Cs matings (Fig. S1, Fig. 1F and G). Thus, the reduction in mating duration arises specifically from Bully males. We therefore chose to use Cs females as a standard background to assess the consequences of male-specific selection for aggression on reproductive behaviors.

(2) Lifespan analysis is done on male progeny of Cs females mated to either genetically more distant Bully or co-evolved Cs males; the longer lifespan and performance on the former is interpreted as a trade-off with aggressiveness, rather than a simple explanation of hybrid vigor.

We appreciate this comment, which again stems from a poor explanation from our part about the origin of the Bully line in the original manuscript. The Bully flies were derived from the same original population as the Cs line. Hybrid vigor typically arises when crossing individuals from distinct populations, which is not the case here as both Bully and CS come from the same population.

To further support our conclusions, we conducted additional experiments using progeny from within-line crosses (Bully males × Bully females) and results revealed the same phenotype: the progeny of these flies also exhibited significantly longer lifespans than Cs males x Cs females progeny. This finding argues against hybrid vigor as the main explanation for the observed phenotype, since both the Bully and Cs crosses result in inbreeding, yet give longer lifespan in Bully. We will include these additional longevity data (currently not included in the manuscript) to strengthen our results and reinforce our interpretation.

(3) Differences in CHCs between Bully and Cs males and Cs females mated to those males are not shown to cause differences in measured behavioral outcomes.

We thank the reviewer for raising this important point regarding causality. One way to establish a causal link between differences in CHCs observed in Bully and Cs flies and the corresponding behavioral outcomes would be to experimentally manipulate CHC profiles. For instance, one could perfume oenocyte-less males with the compounds found in higher abundance in Bully flies, then perform behavioral assays to assess causality. We agree that such experiments would be highly informative in determining the functional roles of specific CHCs elevated in Bully males. However, this approach is technically challenging, as the perfuming technique must be optimized to transfer precise amounts of each compound. For example, this method can be used to gradually perfume flies to assess dose–response behavioral effects, whereas matching exactly the natural concentrations found in individuals, especially given inter-individual variability, remains difficult.

We considered conducting such experiments during our study but did not pursue them for these technical reasons. Nevertheless, we can include a statement in the Discussion acknowledging this as an important future direction to test the causal relationship between CHC variation and behavior.

Reviewer #2 (Public review):

Summary:

The authors compare "Bully" lines, selected for male aggression, to Canton-S controls and find that Bully males have lower mating success, shorter mating durations, and remate sooner. Chemical analyses show Bully males have distinct cuticular hydrocarbons (CHC) signatures and transfer markedly less cVA to females, offering a plausible mechanistic link to weaker mate-guarding.

Paradoxically, Bully males live longer and remain fertile at older ages when CS males no longer mate, indicating a shift in the reproduction-survival trade-off in aggression-selected populations.

Importantly, the work sheds light on proximate mechanisms, demonstrating that shifts in CHCs and pheromone transfer co-occur with changes in fitness traits, thus offering new entry points for understanding life-history evolution.

We thank the reviewer for this positive summary of our work.

Strengths:

The manuscript's strengths lie in its comprehensive and integrative approach framed within an evolutionary context. By combining behavioral assays, chemical profiling, and lifespan measurements, the authors reveal a coherent pattern linking aggression selection to life-history trade-offs. The direct quantification of cVA in female reproductive tracts after mating provides a particularly compelling mechanistic correlate, strengthening the link between behavior and chemical signaling. Findings on altered 5-T and 5-P levels further highlight how chemical communication shapes mating and mate-guarding strategies. Analytical approaches are largely rigorous, and the results provide valuable insights into the pleiotropic effects of selection on socially relevant traits. The study will be of interest to Drosophila biologists working on sexual selection, behavioral evolution, and aging.

We thank the reviewer for recognizing the integrative design and mechanistic contributions of our study.

Weaknesses:

The weaknesses are primarily conceptual rather than procedural. The generality of the findings is uncertain, as selection appears to be represented by only one (and a second closely related) Bully line, limiting conclusions about selection responses versus line-specific drift or founder effects. The causal link between aggression selection and increased longevity is not established: the data show a correlated shift but do not identify mechanisms underlying lifespan extension. In several places, the manuscript uses causal language (e.g., that selection 'influences' longevity or mating strategy) where association would be more accurate; this should be toned down to avoid overstatement. Ecological relevance is also not addressed, since laboratory conditions may bias the balance between costs and benefits of aggression compared with variable natural environments. Addressing these points would strengthen both the impact and clarity of the study.

(1) Generality of findings and potential line effects

We agree that our results presented in the main figures of the manuscript relied mainly on one Bully line (Bully A). To address potential line-specific effects, we replicated key courtship experiments with another independent line, Bully B, selected in parallel from the same Canton-S stock but through distinct selection replicates. The results obtained from Bully B closely matched those from Bully A, suggesting that the observed phenotypes are consistent consequences of aggression selection rather than random drift or founder effects.

(2) Causality versus correlation

We concur that some sentences in the manuscript could overstate causal interpretations. We will revise the text to clearly distinguish correlation from causation and to avoid implying direct causal relationships where data only support association.

(3) Ecological relevance

We appreciate this point. Our experiments were performed under controlled laboratory conditions, which may not fully capture the ecological contexts shaping the costs and benefits of aggression. We will acknowledge this limitation and expand the Discussion to consider how environmental variability could modulate the fitness trade-offs associated with aggression in natural populations.

We thank both reviewers for their constructive feedback, which will help us strengthen the rigor and clarity of the manuscript. We believe that the additional results and revisions will satisfactorily address their concerns.

Recommendations for the authors:

Reviewing Editor Comments:

The major weaknesses raised by the reviewers, namely the flies used in the study (CsxBully compared to CsxCs) where the effect on lifespan could be explained by hybrid vigor and the use of only one Bully and one Cs line that does not allow to link unambiguously the observed effect to the selection for aggression, should be addressed by a different experimental design and additional lines to exclude the effect of non-aggression related factors.

We thank the Reviewing Editor for these comments.

(i) Experimental design and hybrid vigor:

Hybrid vigor typically arises from crosses between genetically divergent populations. In our study, Bully lines were derived from Canton-S background and are not therefore not genetically distant from controls. To directly address this concern, we included new data from Bully × Bully pairs (Figure 1), using independently selected Bully lines. These experiments reproduce the key aggression and courtship phenotypes observed in Cs × Bully assays, indicating that the effects are not attributable to hybrid vigor.

(ii) Use of additional selected lines:

We now include data from two independently selected lines (Bully A and Bully B), both derived from Cs, which show consistent behavioral phenotypes. This supports the conclusion that the observed effects are associated with selection for aggression rather than line-specific artifacts. We note that generating such lines is time- and labor-intensive, and only a few laboratories have established aggression-selected lines in Drosophila melanogaster (e.g., Penn et al., 2010; Dierick et al., 2006; Edwards et al., 2006). Accordingly, we have revised the manuscript to explicitly acknowledge this limitation and to frame our conclusions in terms of association rather than causation.

Reviewer #1 (Recommendations for the authors):

I can't see any way to interpret the data using CsxCs vs CsxBully comparisons.

We thank the reviewer for this important point. This concern appears to arise from the assumption that Bully and Cs represent genetically distinct or non-coevolved populations. However, Bully lines were directly derived from Cs and therefore share a common genetic background. We have clarified this point in the Introduction (lines 104-106) and Results (lines 132-137).

Importantly, we now include additional data showing that key phenotypes, including reduced mating duration, are also observed in Bully × Bully pairings and across independently selected Bully lines (new Figure 1). These results demonstrate that the observed effects are driven by the male genotype and do not depend on the female background.

Because selection for aggression was applied specifically to males, we used Cs females as a standardized background to isolate male-specific effects while minimizing variability arising from female genotype or behavior. This rationale is now explicitly stated in the Results (lines 163-166) and at the beginning of the Discussion (lines 326-330). This experimental design allows interpretation of male-specific effects, and the observed differences cannot be attributed to cross design artifacts or hybrid vigor.

Reviewer #2 (Recommendations for the authors):

Major comments:

(1) Several passages currently imply causality, whereas the data support correlations between selection and trait differences rather than direct causation. This overstatement also appears in section subheadings within the Results, such as "Hyper-aggressive males display reduced mate-guarding efficiency, without compromising female fertility." Please consider toning down the wording by replacing active causal verbs with more neutral phrasing. Additionally, it would be important to include a clear, explicit sentence in the Discussion acknowledging this caveat, as the existing phrase "is associated with changes in reproductive traits" does not fully convey this nuance.

We thank the reviewer for this important comment. We have revised the manuscript throughout, including Results subheadings, to replace causal language with association-based phrasing. We also rephrase the first sentence of the Discussion to clarify that our conclusions are correlational (see line 320).

(2) Figure 1 would benefit from a simple schematic of the behavioral paradigm and the arena, since the authors' arena design minimizes manual handling; a cartoon would help readers quickly grasp the assay flow and the conditions under which interactions occur.

We thank the reviewer for this helpful suggestion. We have added a schematic to Figure 1 illustrating the behavioral paradigm and arena design. Additional details are provided in the Materials and Methods (Trannoy et al., 2015). This improves clarity and accessibility of the experimental design.

(3) In Figures 2A-B and A'-B', the higher post-mating UWE in Bully males is intriguing, but these panels do not actually measure the refractory period. It would be helpful to include 'latency' in the first UWE after mating in these swapped-female conditions. This could also be repeated with pheromone-standardized (cVA/CHC-equalized) decapitated females to disentangle effects of female pheromone load from male sensory perception. In addition, a baseline courtship control (naive males with decapitated virgins) is necessary to test whether Bully males simply have a lower threshold for initiating courtship.

We thank the reviewer for this suggestion. The referenced panels are now shown in Figure 3. We quantified post-mating courtship latency; however, latencies were very short across conditions, and no differences were observed between genotypes. We therefore revised the text to interpret these results in terms of post-mating courtship motivation rather than refractory period. The baseline courtship control with decapitated virgins is provided in Fig 2G. These changes clarify the interpretation of post-mating behavior and address the reviewer’s concerns.

(4) Related to my above point, the results in Figure 2B-B′ raise the possibility that Bully males have reduced perception or neural sensitivity to anti-aphrodisiac pheromones deposited by CS males, which could account for their elevated post-mating courtship; the authors might consider experiments that directly test male sensory responsiveness to these cues or mention this possibility in the Discussion.

We thank the reviewer for this point. We performed additional assays to test males’ sensory responsiveness using binary choice assays and measured the time spent performing UWE towards decapitated females versus males. These results were added in Figure 3-Figure Supp 1, and indicate that both Cs and Bully males displayed courtship preferentially towards females, providing a control for sensory perception.

(5) In multiple figure panels, virgin and mated females are depicted with the same symbols, which makes interpretation confusing.

Thank you for pointing this out. We have updated the figure panels to use distinct symbols for virgin and mated females to improve clarity.

(6) For Figure 4, it would be helpful to provide standalone KM curves for Bully versus CS males, including a separate panel for isolated (never-mated) males, and present mating counts in a separate panel while reporting survival models that incorporate mating frequency (or use it as a time-dependent covariate). Although Figures 4C-D report median survivals, full KM plots and an isolated-male curve are important since mating itself elevates mortality and can otherwise confound intrinsic lifespan differences.

Thank you for this important point to improve clarity of this figure. We have reorganized this figure (now Figure 5) to now, present survival curves first (isolated and group-housed males), followed by lifetime mating counts. This reorganization separates survival from mating activity and addresses the potential confounding effect of mating on lifespan. For the lifetime mating data, we used bar plots rather than curves to better visualize individual mating events.

(7) The following sentences overstate the results and imply causality; consider toning down: "These findings suggest that 5-P and 5-T might contribute to promoting remating in females that have previously mated with Bully males (Figure 2F' and G'). Given that Bully males also showed higher levels of both 5-P and 5-T compared to naïve Cs males (Figure 3B), it is likely that the elevated levels of these compounds observed in females result from their transfer during mating."

We thank the reviewer for this important point. We have rephrased these sentences to remove causal language and instead describe associations between CHC profiles and behavioral outcomes. In particular, statements implying that 5-P and 5-T promote remating or are directly transferred during mating have been revised to reflect correlational evidence only (see lines 253-256).

(8) It is not entirely clear how aggression was quantified in each generation, what proportion of males were selected to breed, and whether the findings generalize beyond a single Bully line (Figure Supplement 1 shows data from a closely related Bully line). Without independent replicate lines or sham-selected controls, it remains difficult to rule out drift or line-specific artifacts, and this limitation should be explicitly acknowledged.

We thank the reviewer for this important point. We have clarified the aggression selection procedure by adding methodological details from Penn et al., including how aggression was quantified and how breeders were selected (lines 104-106 and 131-137). Briefly, independent selection replicates were initiated from the same Canton-S population, generating three lines (Bully A, B, and C), which were maintained separately.

To address generality, we now include data from multiple lines. In particular, a new Figure 1 presents aggression and courtship phenotypes across Bully A, B, and C, and key behavioral results are consistent across independent lines.

We acknowledge that additional independent lines would further strengthen generality; this limitation is now explicitly stated in the Discussion (lines 325-326).

These additions clarify the selection procedure and support that the observed phenotypes are associated with aggression selection rather than line-specific artifacts.

Minor Comments:

(1) Exact sample sizes for every experiment should be included in the main figure legends.

Thank you. We have added the number of replicates in each figure legends.

(2) In Figure 1-Supplement 1, the orientation for depicting mating success is reversed compared to Figure 1, which is a bit jarring; it would be clearer to keep the orientation consistent with the main figure.

Thank you. We have incorporated the results initially presented in Figure 1-Sup 1 into a new Figure 1 with additional results, and have taken into account reviewers’ comment.

(3) For multivariate analyses, I suggest including important details such as group sample sizes, p-value, and the percent variance, etc., in the figure legend rather than keeping this only in Supplementary Table S1.

We have inserted these details directly into the figure legends for clarity.

(4) Why was the food cup used for arenas where decapitated virgins were used in mating assays?

Thank you for pointing this. We now have clarified the experimental procedure in the M&M of the revised manuscript (lines 465-469).

(5) For cartoons in Figure 2, the current yellow background makes it very difficult to distinguish flies drawn in yellow or green. Please adjust to a higher-contrast background or add darker outlines so that the cartoons are clearly legible.

Thank you. We have increased the contrast of the female bodies to ensure the cartoons are clearly distinguishable (now figure 3).

(6) Addition of line numbers in the manuscript would be helpful during the review process.

Line numbers have been added throughout the manuscript.

(7) I noticed a few typos in the manuscript. For example, in the Introduction, "seminal fuids" should be corrected to "seminal fluid." In the Discussion, the phrase "CHCs profiles compared those" requires a "to" before "those." Please carefully review the manuscript for similar errors.

Thank you for pointing this out. We carefully reviewed the manuscript for typos and corrected all identified errors.

Version 2

Abstract

Life history traits are all essential traits linked to development, survival and reproduction, directly influencing individual fitness. Limited environmental or physiological resources force organisms to balance competing demands, leading to fundamental trade-offs, particularly between survival and reproduction. Male aggression is often assumed to enhance reproductive success, yet its broader association with other fitness-related traits remains unclear. To address this, we used Drosophila melanogaster Bully lines selected for high levels of male aggression, to examine relationships between aggression, chemical signaling, and essential life-history traits including reproductive behaviors and lifespan. Our results reveal that increased male aggressiveness shifts the balance of a major life-history trade-off, favoring survival over reproductive success. Bully males exhibited lower mating success, shorter mating duration and less effective chemical mate-guarding. However, this was traded-off with increased lifespan, enabling more reproductive opportunities. We also uncover potential mechanistic underpinnings of the reproductive deficits. Bully males showed differences in their cuticular hydrocarbon (CHCs) profiles and transferred lower levels of cVA to females, a mate-guarding pheromone, thereby weakening their post-mating strategy. Overall, our findings indicate that selection for male aggression is associated with a shift in a key life-history trade-off, highlighting aggression as an evolutionary force shaping adaptation of life-history traits and providing a foundation for future genetic and mechanistic studies.

Introduction

Life history traits, such as development time, fertility, survival, and reproductive success, are key phenotypic components that directly influence individual fitness (Braendle and Paaby, 2024; Roff, 1992; Stearns, 1992). These traits often face trade-offs: investing in one can compromise another (Fisher, 1930; Stearns, 1989; Williams, 1966). For instance, high reproductive effort may reduce survival due to the energetic costs of gamete production or increased mortality associated with mating behaviors or competition (Flatt, 2011; 2020). Trade-offs also emerge between pre- and post-copulatory strategies (Simmons, 2001; Warner et al., 1995). Males prioritizing pre-copulatory investment, such as aggressive competition for mates, may have limited resources for post-copulatory investment linked to sperm competition such as mate guarding (Andersson, 1994; Brennan and Orbach, 2021). Understanding how such trade-offs are shaped by natural selection offers insight into how organisms optimize fitness in competitive social environments (Roff, 1992; Stearns, 1976).

The fruit fly Drosophila melanogaster (D. mel.), provides a powerful model for studying life-history trade-offs, with well-characterized developmental, behavioral, and reproductive traits (Flatt, 2020). These traits are shaped by both genetic and environmental factors and can evolve under ecological pressures. Faster development, earlier reproduction and high mating rate, for instance, are advantageous in changing environments but come at the cost of reduced longevity and fecundity (Partridge and Farquhar, 1981; Stearns, 2000). Experimental manipulations of environmental conditions such as temperature and food availability reveal how external factors shape trade-offs between survival and reproduction (Chippindale et al., 1996; Marshall and Sinclair, 2010).

Beyond development and fecundity, D. mel. serves as a model to study a wide range of social behaviors, including aggregation, aggression, and courtship, that are largely mediated by chemical communication (Kim et al., 2017; Kohl et al., 2015). Cuticular hydrocarbons (CHCs) such as some pheromones play diverse roles in social interaction: 7-Tricosene (7-T), a male-specific compound, increases female receptivity and modulates aggression (Grillet et al., 2006; Wang et al., 2011); cis-vaccenyl acetate (cVA), a male-transferred pheromone, attracts females, repels other males, and promotes aggregation (Kurtovic et al., 2007; Laturney and Billeter, 2016); and the female-specific hydrocarbon 7,11-heptacosadiene (7,11-HD) strongly elicits male courtship and supports species-specific mate recognition (Toda et al., 2012).

Among these interactions, male-male competition plays a central role in shaping reproductive strategies. Pre-copulatory competition involves resource defense, courtship displays, and aggressive interactions to secure mating opportunities. Courtship in D.mel. includes stereotyped behaviors such as orientation, wing extensions (producing courtship song), tapping, licking, and attempted copulation (Yamamoto and Koganezawa, 2013). Successful courtship displays lead to female acceptance, outcompeting other males that fail to meet the female’s preferences (Baxter et al., 2018; Hindmarsh Sten et al., 2025b). Aggression intensifies this competition: more aggressive or dominant males often monopolize access to females (Dow and von Schilcher, 1975; Filice and Dukas, 2019; Gao et al., 2024; Nandy et al., 2016; Prunier and Trannoy, 2024). Aggressive behaviors range from wing threats and lunging to high-intensity physical confrontations like tussling and boxing (Chen et al., 2002). Beyond aggressive patterns, males also produce agonistic wing flicks when competing for access to females, which can interfere with a female’s auditory perception of rival males and thereby influence mate choice (Hindmarsh Sten et al., 2025a). These behaviors promote territorial dominance and enhance mating success (Gao et al., 2024), although subordinate males may also achieve reproductive success through first access to females and prolonged copulation (Filice and Dukas, 2019).

Post-copulatory competition further shapes male reproductive success. After mating, males use multiple strategies to ensure paternity, including the transfer of seminal fluid proteins (SFPs) including Sex Peptide (SP) and Accessory gland proteins (Acps), which promote sperm storage, stimulate oviposition, and reduce female receptivity to further mating (Chapman et al., 2003; Fricke et al., 2009). These factors also contribute to the formation of a mating plug that retains sperm, acting as a passive mate-guarding strategy (Avila et al., 2011; Dunham and Rudolf, 2009). In parallel, chemical mate guarding via pheromone transfer (e.g. cVA and 7-T) reduce female attractiveness and their likelihood of re-mating (Billeter et al., 2009; Gillott, 2003; Laturney and Billeter, 2016; Miyamoto and Amrein, 2008; Scott, 1986; Zawistowski and Richmond, 1986). Male reproductive behavior is plastic: in the presence of rivals, males extend their mating duration (Bretman et al., 2009; Bretman et al., 2013; Kim et al., 2013), increase the number of sperm in their ejaculate (Garbaczewska et al., 2013), adapt seminal fuid content in their ejaculate (Hopkins et al., 2019), and present higher levels of aggression toward opponents to guard their mates (Baxter et al., 2015; Bretman et al., 2011; Dore et al., 2021; Mazzi et al., 2009). Together, this demonstrates that male-male aggression influences access to mates and reproductive outcomes through both pre- and post-copulatory competition.

While aggression can enhance male mating success in natural settings, the relationship between aggression and reproductive fitness is not always straightforward. In a previous study, Penn et al. (Penn et al., 2010) generated “Bully” flies through 37 generations of selection for aggression starting from a Canton-S (Cs) population. In the first step, Cs winner males from aggressive encounters were allowed to mate with Cs females. In the resulting F1 progeny, males displaying higher levels of aggression were again selected and mated with their female siblings. The resulted hyper-aggressive males frequently escalated to boxing, the most intense aggression phenotype (Chowdhury et al., 2017), but showed reduced reproductive success with virgin females compared to unselected control males (Penn et al., 2010). This suggests that although aggression may enhance competitive ability, it can also incur fitness costs. This raises important questions about the role of aggression in shaping life-history trade-offs, such as: to what extent does aggression influence life history traits related to reproduction, and does it affect the balance reproduction vs survival?

To address these questions, we investigated whether D. mel. males from Bully lines exhibit changes in life-history trade-offs compared to the unselected starting wild-type flies (Cs lines). We found that hyper-aggressive Bully males show reduced reproductive success and shorter mating duration across their lifespan, indicating lower pre-copulatory success. Their mates present higher remating rate and increased attractiveness, suggesting that mate-guarding strategy of Bully males is less efficient. However, this poor pre- and post-copulatory success was traded-off with an increased lifespan, which is accompanied by late-life reproductive success. We then provided a mechanistic basis for the reduced copulatory success of Bully males. They showed distinct CHCs profiles compared to Cs males, which may contribute to their lower mating success. Moreover, they transferred to females lower levels of cVA, a mate-guarding pheromone, to females, potentially decreasing efficiency of their post-mating strategy. Our findings underscore the complex interplay between aggression, reproductive behavior, survival, and chemical signaling, providing insights into the evolutionary dynamics of life-history trade-offs in competitive social contexts.

Results

Males exhibiting higher aggression intensity also exhibit shorter mating durations

To determine whether selection for male aggressive traits may also be associated with changes in reproductive behaviors, we employed three male lines selected for increased aggression. In a previous study, Penn et al. (Penn et al., 2010) generated the so-called “Bully” flies through 37 generations of artificial selection for male aggression starting from a Canton-S (Cs) population. Briefly, males that won aggressive encounters were allowed to mate with Cs females. In the resulting F1 progeny, males exhibiting higher levels of aggression were again selected and mated with their female siblings. Following the same experimental procedure, several independent selections were performed in parallel, giving rise to the Bully A, Bully B and Bully C lines. These Bully lines were originally derived from Cs flies and do not represent distinct populations per se, but rather result from male-specific selection on aggressive behavior. Males from these three lines (Bully A, B, and C) were tested in standardized aggression assays (Trannoy et al., 2015b) (Figure 1A). Across same-line contests, males from all three Bully lines did not differ from Cs males in latency to lunge (Figure 1B), but displayed significant increase in the number of lunges (Figure 1C), the number of boxing events (Figure 1D), and boxing frequency (defined as the proportion of male pairs displaying at least one boxing event during the 10-min observation period) (Figure 1E) compared to Cs males. However, the magnitude of this phenotype differed across lines: Bully A and Bully B males exhibited higher aggression levels than Bully C males, which showed an intermediate phenotype (Figure 1E).

Courtship and aggressive behaviors in Cs, Bully A, Bully B and Bully C lines.

(A) Schema of the experimental set up to record aggressive and courtship behaviors. Couples of flies and presence of food cup are mentioned according to the figure of this article. (B) Latency to lunge of Cs, Bully A, Bully B and Bully C males in aggression assay with a territory against a male of the same line (nCs=29, nBullyA=24, nBullyB=12, nBullyC=11, LM and permutation test (Monte-Carlo test): P=0.136). (C) Number of lunges of Cs, Bully A, Bully B and Bully C male pairs in aggression assays (nCs=29, nBullyA=24, nBullyB=12, nBullyC=11, GLM quasi-Poisson and type II analysis of deviance table (Likelihood Ratio Test): P=1.622×10−10, LRχ2=48.556, post-hoc: PCs_BullyA<0.001, z ratio= −4.495, PCs_BullyB<0.001, z.ratio= −5.911, PCs_BullyC=0.018, z.ratio= −2.836, PBullyA_BullyB=0.042, z ratio= −2.308, PBullyA_BullyC=0.232, z ratio= 1.195, PBullyB_BullyC=0.018, z ratio=2.797). (D) Number of boxing events of Cs, Bully A, Bully B and Bully C male pairs in aggression assays (nCs=29, nBullyA=24, nBullyB=12, nBullyC=11, GLM Poisson and type II analysis of deviance table (Likelihood Ratio Test): P=2.2×10−16, LRχ2=2511.9, post-hoc: PCs_BullyA<0.001, z ratio= −6.819, PCs_BullyB<0.001, z.ratio= −6.839, PCs_BullyC<0.001, z.ratio= −3.649, PBullyA_BullyB=0.728, z ratio= −0.347, PBullyA_BullyC<0.001, z ratio=5.392, PBullyB_BullyC<0.001, z ratio=5.366). (E) Non-competitive mating success of Cs, Bully A, Bully B and Bully C males paired with a virgin Cs female (nCs=37, nBullyA=31, nBullyB=33, nBullyC=21, GLM binomial and type II analysis of deviance table (Likelihood Ratio Test): P=0.012, LRχ2=11.040, post-hoc: PCs_BullyA=0.934, z ratio=0.728, PCs_BullyB=0.099, z.ratio=2.395, PCs_BullyC=0.868, z.ratio=1.060, PBullyA_BullyB=0.206, z ratio=2.043, PBullyA_BullyC=0.934, z ratio=0.406, PBullyB_BullyC=0.522, z ratio=-1.513). (F) Proportion of time spent by Cs and Cs, Bully A, Bully B and Bully C males doing Unilateral Wing Extension (UWE) toward a Cs virgin female in non-competitive courtship assays (nCs=37, nBullyA=31, nBullyB=33, nBullyC=21, LM and type II Anova table: P=0.088, F=2.235). (G) Latency to court (time difference between first interaction and first UWE) of Cs, Bully A, Bully B and Bully C males paired with a virgin Cs female (nCs=37, nBullyA=31, nBullyB=33, nBullyC=21, LM and type II Anova table: P=0.014, F=3.707, post-hoc: PCs_BullyA=0.338, t.ratio= −1.740, PCs_BullyB=0.433, t.ratio= 1.469, PCs_BullyC=0.524, t.ratio= −1.127, PBullyA_BullyB=0.015, t.ratio=3.101, PBullyA_BullyC=0.683, t.ratio=0.661, PBullyB_BullyC=0.099, t.ratio= −2.364). (H) Latency to mate (time difference between first interaction and start of the mating) of Cs, Bully A, Bully B and Bully C males paired with a virgin Cs female (nCs=36, nBullyA=29, nBullyB=24, nBullyC=19, LM and type II Anova table: P=8.527×10−4, F=5.970, post-hoc: PCs_BullyA=0.482, t.ratio= −1.413, PCs_BullyB<0.001, t.ratio= −4.191, PCs_BullyC=0.581, t.ratio= −1.062, PBullyA_BullyB=0.038, t.ratio= −2.725, PBullyA_BullyC=0.863, t.ratio=0.173, PBullyB_BullyC=0.041, t.ratio= 2.615). (I) Mating duration of Cs, Bully A, Bully B and Bully C males paired with a virgin Cs female (nCs=36, nBullyA=29, nBullyB=24, nBullyC=19, LM and type II Anova table: P=8.527×10−4, F=5.970, post-hoc: PCs_BullyA=0.482, t.ratio= −1.413, PCs_BullyB<0.001, t.ratio= −4.191, PCs_BullyC=0.581, t.ratio= −1.062, PBullyA_BullyB=0.038, t.ratio= −2.725, PBullyA_BullyC=0.863, t.ratio=0.173, PBullyB_BullyC=0.041, t.ratio= 2.615). For all graphs, different letters mean a statistical difference. Statistical details are given in Figure 1-table supplement 1.

We next assessed courtship and mating behaviors by pairing males and females from the same line and scored courtship metrics. The proportion of time spent by males displaying unilateral wing extension (UWE) and latency to court did not differ significantly between lines (Figure 1F-G), while Bully B males exhibited a longer latency to mate compared to other pairings (Figure 1H). In contrast, mating duration (MD) was significantly reduced in Bully A and Bully B pairings compared to Cs (Figure 1I), while Bully C males again show an intermediate phenotype (Figure 1I).

These results indicate that males from the Bully A and Bully B lines, classified as hyper-aggressive in same-line competition (Figure 1B-E), also exhibit reduced mating duration when paired with same-line females, whereas Bully C males do not (Figure 1I). Because the Bully A line (hereafter named Bully) has been already employed in previous studies to measure differential gene expression profile (Chowdhury et al., 2017; Penn et al., 2010), we chose to conduct subsequent experiments on this line. Overall, these results support our initial hypothesis that selection for aggressive traits may have influenced other behavioral traits linked to reproductive behaviors.

Highly aggressive males exhibit reduced mating success and have shorter mating duration

As the selection for aggression was male-specific (Penn et al., 2010), we next aimed to focus on examining male reproductive behaviors while avoiding confounding effects of female behavioral variation. Thus, all subsequent courtship and mating assays were performed by pairing Cs or Bully A males with Cs females, which served as a standardized female background. We first assessed mating success in competitive assays, where a Cs male and a Bully male competed for access to a single female. Under these conditions, no significant mating advantage was observed for Bully males (Figure 2A). We then assessed mating success in non-competitive assays by pairing each male with a virgin Cs female for 30 minutes. Under these conditions, Bully males showed a significantly reduced probability of mating compared to Cs males (Figure 2B), despite similar levels of courtship behavior, as measured by UWE (Figure 2C), courtship latency (Figure 2D), and mating latency among successful pairs (Figures 2E). We next examined copulation per se and found that MD was significantly shorter in Bully males, with a reduction of approximately 25% compared to Cs males (Figure 2F), consistent with the phenotype observed when males were paired with females from the same line (Figure 1I). Importantly, this reduction in MD persisted when Bully males were paired with decapitated Cs females (Figure 2G), ruling out female-driven termination of mating and demonstrating that the effect is male-specific.

Bully males exhibit reduced mating success and allocate less time to mating.

(A) In competitive courtship assays, the mating success of Cs and Bully males was scored when paired with a single virgin Cs female (nCs=19, nBully=19, GLM binomial and type II analysis of deviance table (Likelihood Ratio): Test, P=0.329, LRχ2=0.951). (B) In non-competitive courtship assays, the mating success of Cs and Bully males paired with a virgin Cs female were analyzed (nCs=64, nBully=41, GLM binomial and type II analysis of deviance table (Likelihood Ratio Test): P=0.037, LRχ2=4.347). (C) Proportion of time spent by Cs and Bully males doing Unilateral Wing Extension (UWE) toward a Cs virgin female in non-competitive courtship assays (nCs=64, nBully=41, Fisher Pitman permutation test: P=0.877, t= −0.131). (D) Latency to court (time difference between first interaction and first UWE) of Cs and Bully males paired with a virgin Cs female (nCs=64, nBully=41, Fisher Pitman permutation test: P=0.319, Z=1.043). (E) Latency to mate (time difference between first interaction and start of the mating) of Cs and Bully males paired with a virgin Cs female (nCs=61, nBully=34, Fisher Pitman permutation test: P=0.266, Z=1.170). (F) Mating duration of Cs and Bully males paired with a virgin Cs female (nCs=61, nBully=34, T-test: P=6.009×10−10, t= −6.917) or (G) a virgin Cs decapitated female (nCs=26, nBully=20, T-test: P=0.022, t= −2.367). (H) Mating duration of Cs and Bully males that were previously raised in group of 10 males or in social isolation (nCs-isolated=22, nCs-group=32, nBully-isolated=23, nBully-group=25, LM and type II ANOVA: F-test, P=7.582×10−11, F=21.669, post-hoc: PCs-isolated_Cs-group=0.013, F=6.583, PBully-isolated_Bully-group=2.532×10−4, F=15.731, PCs-isolated_Bully-isolated=4.263×10−6, F=27.697, PCs-group_Bully-group=1.884×10−4, F=16.032). For all graphs, stars indicate significant differences (* P<0.05, ** P< 0.01, *** P<0.001) and statistical details are given in Figure 1-table supplement 1.

As group-housed males have lengthened MD (Bretman et al., 2013; Kim et al., 2013), we compared MD of singly and group-housed males. We found that both Cs and Bully group-housed males exhibited significantly extended mating durations compared to single-housed males (Figure 1H). However, the MD in Bully males remained shorter compared to Cs males in group-housed condition (Figure 2H), indicating that Cs and Bully males respond equally to early social experience

Together, these results show that Bully males display reduced mating success and shorter copulation duration, indicating differences in pre-copulatory outcomes and copulatory investment. This supports our main hypothesis that selecting males for aggressive traits may also had influenced one or both pre- and post-copulatory strategies.

Post-mating behaviors are affected in hyper-aggressive males

Given these differences in mating performance, we asked whether they compensate through distinct post-mating strategies. To investigate this, we first quantified courtship behavior immediately after copulation by measuring unilateral wing extension (UWE) directed toward the just-mated female. While Cs males performed minimal UWE towards just-mated females (Figure 3A’), Bully males displayed significantly higher levels of post-mating courtship in this context (Figure 3A’). This effect persisted when females were swapped after mating, with Bully males showing increased UWE towards Cs-mated females (Figure 3B–B’). This indicates that this phenotype reflects an intrinsic increase in courtship drive rather than a female-dependent effect. This elevated post-mating motivation led us to hypothesize that Bully males might achieve higher remating rates. To test this, we conducted sequential mating assays in which a single Cs or Bully male was consecutively paired with six virgin females over a 3-hour period and we scored the number of mating. Contrary to our expectation, Bully males did not mate with more females than Cs controls (Figure 3C), despite maintaining consistently shorter MD across successive mating (Figures 3D). As mating accompanies ejaculate transfer and incurs a reduction in sperm and seminal fluid reserve in D. mel. males (Hihara, 1981; Sirot et al., 2009), we further examined whether shorter MD in Bully males could reflect more efficient allocation, allowing fertilization of a greater number of females. However, analysis of progeny production revealed no significant differences between Bully- and Cs-mated females across successive matings (Figure 3E). Thus, increased post-mating courtship in Bully males does not translate into improved reproductive output.

Hyper-aggressive males exhibit changes in mate-guarding features.

(A, B, F, G). Mating duration of Cs and Bully males paired with a virgin Cs female (A: nCs=21, nBully=22, T-test: P=2.912×10−11, t= −8.759, B - nCs=24, nBully=24, T-test: P=5.425×10−7, t= −5.818, F - nCs=25, nBully=25, T-test: P=7.139×10−4, t= −3.617, G - nCs=30, nBully=24, T-test: P=8.743×10−5, t= −4.255). (A’) After mating ended, the proportion of time spent by each male to court performing UWE towards the just mated female was scored for 10 minutes (nCs=21, nBully=22, Fisher Pitman permutation test: P<0.001, Z=3.239). (B’) The proportion of time spent by the males performing UWE an unfamiliar mated female was scored for 10 minutes (nCs=24, nBully=24, Fisher Pitman permutation test: P=0.007, Z= −2.679). (C) Total number of mating reached by Cs and Bully males over six consecutive mating with virgin Cs females (type II analysis of deviance table (nCs=18♂, nBully=18♂, Wald Chi-square test): P=0.659, χ2=0.195). (D) Mating duration of Cs and Bully males over six consecutive mating with virgin Cs females (nCs=18♂, nBully=18♂, LMM and Satterthwaite T-test: Pinteraction=0.001, t= −3.261). A significant interaction between males’ genotype and number of mating was found, indicating that Cs decreased their mating duration with the number of mating, while Bully males did not. (E) Proportion of mating that gave rise to progeny in consecutive mating assay (nCs=18♂, nBully=18♂, GLMM and Wald Chi-square test: Pmale_genotype=0.279, χ2=1.173, Pmating=0.847, χ2=2.019). (F’) Proportion of time spent by naïve Cs and Bully males to display UWE either toward Cs-mated or Bully-mated decapitated female over an observation period of 2 minutes (nCs-Cs-mated=12, nCs-Bully-mated=15, nBully-Cs-mated=9, nBully-Bully-mated=10, LMM and permutation test (Monte-Carlo test): Pmale_genotype<0.001, Pmated_female=0.457). (G’) Proportion of Cs-mated and Bully-mated females that re-mated 12 or 14 days after a first mating (nCs-12days=19, nBully-12days=15, nCs-14days=11, nBully-14days=9, GLM binomial and type II analysis of deviance table (Likelihood Ratio Test): Pmated_female=3.639×10−4, LRχ2=12.708, Pday=0.385, LRχ2=0.756). For graphs A, A’, B, B’, F, F’, G and G’, stars indicate significant differences (* P < 0.05, ** P < 0.01, *** P < 0.001) and statistical details are given in Figure 1-table supplement 1.

We next investigated whether Bully males differ in their ability to modify female post-mating state. During mating, males transfer pheromonal cues, including cuticular hydrocarbons (CHCs), that reduce female attractiveness to rival males. We reasoned that the shorter MD of Bully males might limit the transfer of such mate-guarding signals, thereby reducing the extent of female unattractiveness. To test this independently of female behavior, we performed binary choice assays using decapitated females previously mated with either Cs or Bully males. Both Cs and Bully males spent significantly more time courting females that had mated with Bully males (Figure 3F’, green vs yellow bars), indicating that these females remain more attractive and suggesting reduced transfer of inhibitory cues. To validate the experimental procedure and rule out potential deficits in sensory perception, we confirmed that both Cs and Bully males spent significantly more time performing UWE toward a decapitated female than toward a decapitated male (Figure 3-figure supplement 1). Finally, we assessed the functional consequences of this change in female state on remating behavior. Virgin Cs females were first mated with either Cs or Bully males and subsequently paired with naïve Cs males several days later. Females previously mated with Bully males, remated significantly more than females that had mated with Cs males (Figure 3G’, green), consistent with weaker mate-guarding.

Together, these results indicate that Bully males fail to offset their poor pre-copulatory success with more efficient post-mating strategies. Despite increased post-mating courtship motivation, they do not achieve higher remating rates or reproductive output, and their mates remain more attractive and more likely to remate. This dissociation indicates that increased courtship motivation is not predictive of reproductive success in this context. These findings suggest that selection for aggressive traits in males appears to compromise both pre- and post-copulatory aspects of male reproductive performance.

Hyper-aggressive males present differences in their CHCs profiles and transfer lower quantities of mate-guarding pheromones to females

To investigate mechanisms underlying reduced mating success (Figure 2B) and changes in mate-guarding behaviors (Figure 3F-G) in Bully males, we quantified the CHCs profiles of naïve Cs and Bully males. PCA analysis revealed a clear separation between the two groups, indicating distinct CHCs profiles between naïve Cs and Bully males (Figure 4A). Quantification of individual compounds showed that Bully males exhibited significantly higher levels of alkanes, monoenes, and methyl-alkanes (Figure 4B, Table 1).

Hyper-aggressive males exhibit altered CHCs profiles and transfer reduced amounts of mate-guarding pheromones to females.

(A) PCA analysis on mean concentration for each Cuticular Hydrocarbon (CHCs) compound of naïve Cs and Bully. Their CHCs profiles were found similar (nCs=10♂, nBully=10♂, PerMANOVA: P=0.002, F=6.789, R2=0.274). (B) Relative concentrations of Bully’s CHCs were calculated with respect to those of Cs males. The CHCs levels in Cs males were normalized to 1, and Bully’s CHC concentrations were expressed relative to this reference (nCs=10♂, nBully=10♂, LMM and permutation test (Monte-Carlo test): P=0.009, post-hoc: PAlkanes=0.036, PMonoenes=0.002, PMethyl alkanes=0.048, PnC21<0.001, PC22:1(9)=0.892, PC22:1(7)=0.568, PcVA=0.731, PnC22=0.854, PC23:1(9)=0.175, PC23:1(7)=0.062, PC23:1(5)<0.001, PnC23=0.014, PC24:1(9)=0.345, PC24:1(7<0.001, P24:1(5)=0.015, PnC24=0.002, P2MeC24=0.548, PC25:1(9)=0.524, PC25:1(7)<0.001, PC25:1(5)<0.001, PnC25<0.001, P2MeC26=0.159, PC27:1(7)<0.001, PnC27=0.078, P2MeC28=0.003, PnC29=0.005). (C) PCA analysis on mean concentration for each CHCs compound of virgin, Cs-mated and Bully-mated females (nvirgin=10♀, nCs-mated=16♀, nBully-mated=14♀, PerMANOVA: P=0.017, F=2.594, R2=0.123, post-hoc: Pvirgin_Cs-mated=0.001, F=19.434, R2=0.447, Pvirgin_Bully-mated=0.034, F=2.266, R2=0.093, PCs-mated_Bully-mated=0.746, F=0.428, R2=0.015). (D) Relative concentrations of CHCs compounds in mated females compared to virgin females. CHCs levels in virgin females were normalized to 1, and the concentrations in mated females were expressed relative to this baseline (LMM and permutation test (nvirgin=10♀, nCs-mated=16♀, nBully-mated=14♀, Monte-Carlo test: P=0.618) Even though the p-value was not significant, post-hoc: PAlkanes=0.705, PMonoenes=0.294, PDienes=0.715, PMethyl alkanes=0.750, PnC21=0.158, PnC22=0.927, P7,11C23:2<0.001, P2MeC22<0.001, PC23:1(9)=0.023, PC23:1(7)=0.001, PC23:1(5)=0.016, PnC23=0.908, PnC24=0.899, P7,11C25:2<0.001, P2MeC24<0.001, PC25:1(9)=0.832, PC25:1(7)=0.047, PC25:1(5)<0.001, PnC25=0.769, P9,13C27:2<0.001, P7,11C27:2=0.747, P2MeC26=0.884, PC27:1(9)=0.016, PC27:1(7)=0.625, PC27:1(5)<0.001, PnC27=0.367, P9,13C29:2<0.001, P7,11C23:2=0.899, P2MeC28=0.413, PC29:1(9)<0.001, PC29:1(7)<0.001, PnC29=0.022). Significant differences are indicated by letters: a indicates a significant difference between virgin and Cs-mated females; b indicates a significant difference between virgin and Bully-mated females; and c indicates a significant difference between Cs- and Bully-mated females (a P < 0.05; aa P < 0.01; aaa P < 0.001). (E) Mean ± SE concentration in ng of CHCs compounds that were not present in virgin females (nvirgin=10♀, nCs-mated=16♀, nBully-mated=14♀, PC23:1(6)<0.001, PC25:1(6)<0.001, PcVA=0.001). (F) Mating duration of Cs and Bully males with a Cs female (nCs-mated=17, nBully-mated=14, T-test: P=0.003, t= −3.359). (F’) cVA concentration measured within Cs- and Bully-mated genitalia tract right after mating (nCs-mated=17, nBully-mated=14, Fisher Pitman permutation test: P=0.001, Z= −2.390). For graphs E, F and F’, stars indicate significant differences (* P < 0.05, ** P < 0.01, *** P < 0.001) and statistical details are given in Figure 1-table supplement 1.

We next examined whether mating with Bully males induces changes in CHC profiles of females compared to those that have mated with Cs males. To test this hypothesis, we extracted CHCs from the cuticles of Cs virgin females and Cs females mated to either a Cs or Bully males and compared their CHCs profile. PCA analysis showed that Cs virgin females form a distinct cluster from mated females (Figure 4C), indicating that mating induces substantial changes in female CHC profiles. The first two principal components accounted for most of the variance, and several compounds, including C23:1(6), C25:1(6), and cVA, contributed strongly to this separation (Figure 4-figure supplement 1). Consistent with this, individual compound analysis revealed that male-specific CHCs, such as 7-tricosene (7-T), 9-tricosene (9-T), and cVA, were significantly increased in mated females compared to virgins (Figure 4D–E, Table 2). These results are consistent with previous studies showing that 7-T and 9-T are acquired by females during mating (Everaerts et al., 2010; Laturney and Billeter, 2016).

We then compared CHC profiles between females mated with Cs or Bully males. While both groups exhibited the expected increase in male-derived compounds, females mated with Bully males showed significantly higher levels of 5-tricosene (5-T) and 5-pentacosene (5-P) (Figure 4D, Table 2). As Bully males also showed higher levels of both 5-P and 5-T compared to naïve Cs males (Figure 4B), it is possible that the elevated levels of these compounds observed in females could result from their transfer during mating.

cVA, a male-specific compound that acts as an anti-aphrodisiac and a mate-guarding molecule, is primarily transferred to the female’s genital tract during mating, with only trace amounts deposited on the cuticle (Laturney and Billeter, 2016). Females with higher cVA levels (Kurtovic et al., 2007), combined with 7-T, are less attractive to males (Laturney and Billeter, 2016). Given that remating rate of Bully-mated females was higher than that of Cs-mated females (Figure 3G’), we hypothesized that Bully-mated females would receive less cVA in their genital tract during mating, thus decreasing their attractiveness than when mating with Cs males. To test this, we conducted mating assays with Cs females paired with either Cs or Bully males, followed by dissection of the females’ genital tracts to quantify the amount of cVA transferred. We found that females mated with Bully males had about three times lower cVA quantities in their genital tracts compared to those mated with Cs males (Figure 4F’). Combined with higher quantities of 5-T and 5-P, reduced cVA level in genital tract likely contribute to the increased attractiveness and higher remating rates observed in females that previously mated with Bully males.

Overall, our findings demonstrate that naïve Bully males differ from Cs males in their CHC profiles and transfer a distinct quantity of chemical compounds to females during mating. These observations are consistent with a model in which changes in chemical signaling in Bully males is associated with both reduced mating success and weaker mate-guarding-related behaviors. In particular, reduced cVA transfer, combined with elevated levels of specific hydrocarbons such as 5-T and 5-P in mated females, may contribute to the increased attractiveness and remating propensity observed in females that previously mated with Bully males.

Reduced copulatory investment in aggressive males is associated with extended lifespan

Based on the main life-history trade-off between reproduction and survival, we next examined whether reduced pre- and post-copulatory investment observed in Bully males would be accompanied by increased longevity. To this end, we conducted longevity assays and monitored lifespan in Cs and Bully males. When housed individually, Bully males survived significantly longer than Cs males (Figure 5A, green vs yellow curves). Median survival (EC50) was increased from 57 days in Cs males to 82 days in Bully males (Figure 5A), and maximum lifespan was also extended from 90 days in Cs males to 100 days in Bully (Figure 5A). To assess whether this difference could reflect hybrid vigor, we measured lifespan in heterozygous Bully males generated by crossing Cs females with Bully males. The survival of these heterozygous males was comparable to that of homozygous Bully males (Figure 5A, red curve), with males showing an EC50 of 79 days and a maximum lifespan of 106 days. This result indicates that Bully background positively influence males’ lifespan and argue against hybrid vigor as the main explanation for the observed phenotype. We next evaluated survival under more naturalistic social conditions by maintaining males in mixed-sex groups. Under these conditions, Bully males again showed increased survival compared to Cs males, with both higher median lifespan and extended maximum lifespan (Figure 5B). Bully males had a median survival of 56 days and a maximum lifespan of 71 days, compared to Cs males, which had a median survival of 43 days and all died by day 67 (Figure 5B).

Aggressive males offset their reduced reproductive success with a longer lifespan and preserved reproductive success in later life.

(A) Survival curves of Cs (yellow) and homozygous (green) and heterozygous (red) Bully males (Weibull regression model (n =50, n =50, n =51, Likelihood Ratio Test): χ2=23.91, p=2.6×10−5, post-hoc: PCs – Bully_homo<0.001, t.ratio= 4.061, PCs – Bully_hetero< 0.001, t.ratio=5.088, PBully_homo – Bully_hetero=0.318, t.ratio = −1.001). Cs males had a median survival (EC50) of 57 days, with the last individual dying at day 90 (yellow), while progeny from homozygous Bully males showed an EC50 of 82 days, with the final death at day 101 (green) and heterozygous Bully males showed an EC50 of 79 days, with the final death at day 106 (red). (B) Survival curves of group of 10 Cs and 10 homozygous Bully males raised in group with 12 Cs females. Bully males had a median survival (EC50) of 56 days, with the last individual dying at day 71 (red line), while Cs males showed an EC50 of 43 days, with the final death at day 67 (yellow line) (nCs=50, nBully-hetero=50, Cox proportional hazards regression model with stratification by group (Likelihood Ratio Test): P=7×10−5, χ2=15.94). Statistical details are given in Figure 1-table supplement. (C) Mating duration of Cs and Bully males with virgin Cs females across their lifespan (nCs-J0=24, nBully-J14=23, LMM and Satterthwaite T-test: Pinteraction =3.360×10−5, t=4.227). A significant interaction between male genotype and age was observed: Bully males maintained a stable mating duration over time, whereas Cs males showed an increase in mating duration with successive mating events. (D) Number of mating throughout the lifespan of Cs and Bully males (nCs-J0=24, nBully-J14=23, GLM binomial and Wald Z-test: Pinteraction=0.039, Z= −2.068). Numbers in bars indicate the proportion of mating. Arrows denote the last mating event and the number of offspring produced. Statistical details are given in Figure 1-table supplement 1. Table 1: Comparison of CHCs concentrations between naïve Cs and Bully males. Mean ± SE concentrations are provided in ng for each compound and each group of compounds. In statistics columns, stars represent significant differences (* P < 0.05, ** P < 0.01, *** P < 0.001). Statistical details are given in Figure 1-table supplement 1.

We then assessed how reproductive performance varies across the lifespan. Individual males were paired weekly with virgin Cs females until death, and mating success and mating duration were recorded. Throughout their lifespan, Bully males consistently displayed shorter MD than Cs males (Figure 5C). While mating duration increased with age in Cs males, a pattern reported in another Drosophila species (Dhole and Pfennig, 2014), this effect was not observed in Bully males, whose MD remained stable across mating and throughout their lives (Figure 5C, green). Analysis of individual male slopes confirmed this divergence: most Cs males showed increasing MD across mating events, while Bully males did not (Figure 5-figure supplement 1). Mating success followed a similar age-dependent pattern. Bully males showed reduced mating success compared to Cs males during early and mid-adulthood (until 70 days), but continued to mate at ages when Cs males had ceased reproductive activity (Figure 5D). While all Cs males stopped mating by day 78 and died by day 97 (Figure 5D, yellow), Bully males maintained mating activity until day 98 and survived up to 119 days (Figure 5D, green). In addition, late-life mating involving 98-days-old Bully males produced viable offspring (Figure 5D, green), demonstrating that Bully males remain fertile at older ages.

Together, our findings demonstrate that Bully males exhibit both reduced copulatory investment and extended lifespan, as well as a shift in the temporal pattern of reproductive activity. These observations are consistent with a scenario in which reduced investment in mating success and mating duration is associated with extended lifespan and sustained reproductive output later in life. Rather than enhancing early-life reproductive success, Bully males maintain reproductive capacity over a longer time window, suggesting a redistribution of reproductive effort across the lifespan.

Discussion

Our study reveals that selection for aggressive traits in D. mel. males correlates with changes in key life-history trade-off, in which survival is extended while reproductive success is reduced. This pattern challenges the hypothesis that aggression necessarily enhances mating success (Andersson, 1994; West-Eberhard, 1983), and indicates that aggressive phenotypes may incur substantial reproductive costs. Consistent phenotypes across independent Bully lines support a link between aggression selection and the reproductive traits described here. However, the absence of additional selected lines limits our ability to fully exclude contributions of genetic drift or line-specific effects. Because selection for aggression was applied specifically to males, we used a standardized female background (Cs) to isolate male-driven effects. While this design minimizes female variability, it does not fully disentangle male and female genotype interactions. However, the persistence of key phenotypes in Bully × Bully pairings and across independent selected lines supports a primary contribution of male genotype. Aggression therefore does not act as an isolated fitness-enhancing trait, but rather as one component of a coordinated life-history strategy in which gains in competitive ability are coupled to constraints on reproductive performance. The increased survival of highly aggressive males further raises important questions regarding the evolutionary consequences of aggression. While Bully males have reduced mating success compared to Cs males, their extended lifespan may allow them to sustain mating activity over longer periods and produce viable offspring at later ages, challenging the classical view that early reproductive success is the primary driver of male fitness (Flatt, 2011). This shift suggests that alternative reproductive strategies, even those associated with lower mating success and mate-guarding, can persist if they confer long-term survival advantages under laboratories conditions.

At the behavioral level, highly aggressive Bully A males exhibit changes in both pre- and post-copulatory phases. They achieve lower mating success and engage in shorter copulations than less aggressive Cs males, suggesting reduced efficiency in mate acquisition and potentially in sperm transfer. Despite higher courtship motivation after mating, Bully males do not achieve higher remating success. Instead, females previously mated with these males remate more readily, indicating reduced mate-guarding efficiency and possibly increased attractiveness. These observations point to a decoupling between courtship intensity and mating outcome, indicating that increased behavioral drive does not translate into reproductive advantage. While our study did not show a causal link, our results suggest that artificial selection applied to male aggressive traits to produce the Bully A line, is associated with changes in reproductive traits, enhanced survival, and increased CHCs levels.

These behavioral differences are closely associated with significant modifications in chemical signaling. Cuticular hydrocarbons (CHCs), which are crucial to mate recognition and sexual communication (Billeter and Wolfner, 2018; Yew and Chung, 2017) differ significantly between Bully A and Cs males. We show that naïve highly aggressive individuals present distinct chemical profiles compared to unselected wild-type flies, with higher levels of most CHCs. Importantly, CHCs also contribute to physiological functions such as protection against desiccation (Nayal et al., 2024), suggesting that changes in CHC composition may simultaneously influence both reproductive performances and survival. Thus, selection on aggressive traits may have been accompanied by modifications in overall chemical signaling that would impact both physiological resilience and social behaviors.

Although higher CHC level might be expected to enhance male attractiveness, Bully males display reduced mating success, suggesting a non-linear or dose-dependent relationship between CHC levels and female acceptance. Female mate choice relies on finely tuned blends of chemical cues, and changes in these optimal ratios may impair signal interpretation (Billeter and Wolfner, 2018). In this context, excessive levels of specific CHCs could reduce courtship efficiency rather than enhance it. In addition to major compounds such as 7-tricosene, male attractiveness and courtship performance are influenced by other CHC, including palmitoleic acid, which signals through the Or47b receptor and modulates courtship behavior in an age-dependent manner (Dweck et al., 2015; Kohlmeier et al., 2021; Lin et al., 2016). This pathway enables older males to initiate courtship more rapidly and with greater intensity, potentially compensating for age-related declines in competitiveness. Although palmitoleic acid levels were not measured here, change in level of such compounds could further contribute to decreasing mating success despite overall higher CHC level.

Modifications in chemical communication extend into the post-mating context, where they potentially intersect with male reproductive performance. Females mated with Bully males display distinct CHC profiles, with higher levels of 5-tricosene (5-T) and 5-pentacosene (5-P), together with reduced transfer of cis-vaccenyl acetate (cVA) into the female genital tract, which may fail to efficiently decrease their attractiveness, thus triggering higher remating rate. This interpretation is consistent with previous studies showing that lower cVA levels decrease the effectiveness of chemical mate guarding and accelerate the recovery of female attractiveness (Ejima, 2015). At the same time, the behavioral effects of cVA depend on its interaction with other CHCs (Laturney and Billeter, 2016; Verschut et al., 2023), suggesting that changes in the overall chemical profile may further modulate how mated females are perceived by subsequent males. These results indicate that selection on aggressive traits can indirectly shape post-copulatory sexual selection by altering female chemical signatures. While causal manipulation of CHC profiles would be required to directly test their functional role, the consistent association between CHC composition, cVA transfer, and behavioral outcomes across assays of our study identifies chemical signaling as a strong candidate mechanism.

Taken together, these observations support the idea that cuticular hydrocarbons constitute a central mechanistic substrate linking aggression, reproduction, and survival. Rather than hypothezising selection on multiple behavioral traits, the phenotypic differences observed in our study may alternatively arise from selection acting on genes involved in chemical signaling. Because CHCs influence both social communication and physiological resilience (Nayal et al., 2024), selection that have acted on CHC production or emission would be expected to generate correlated effects across social behavior, reproductive success, and survival, and physiological resilience, thereby producing the unique life-history pattern observed in Bully males.

At the genetic level, this integration may be supported by pleiotropic factors or by the co-selection of partially independent loci affecting multiple traits. In this study, we mainly focused on Bully A males, as transcriptomic analyses have identified a limited set of genes differentially expressed between Bully A and Cs males, including CG13646, a putative transmembrane transporter whose down-regulation increases aggressive behavior (Chowdhury et al., 2017). These candidate genes provide an entry point for dissecting the molecular and genetic logic underlying the observed phenotypes, particularly in determining whether changes in aggression, CHC composition, survival and life-history traits arise from common genetic regulation.

In summary, our findings indicate that selection for male aggression is associated with a coordinated shift in reproductive investment, chemical signaling and lifespan, consistent with a redistribution of reproductive effort across the lifespan. This highlights aggression as a potential entry point into the coordinated evolution of reproductive and physiological traits.

Material and methods

Experimental model and subject details

Fly stocks

Drosophila melanogaster strains were raised at 25°C under a 12h:12h light/dark cycle (LD= 9:00a.m. – 9:00p.m.) on standard fly food (for 1l of water, 74g corn flour, 28g yeast, 40g sugar, 7.5g agar, 10ml Moldex). Canton-S (Cs) flies (from Edward Kravitz’s laboratory, RRID:BDSC_64349) were used as wild-type. The Bully A, B and C lines were originally selected in Edward Kravitz’s laboratory from Cs during 37 generations (Penn et al., 2010), resulting in hyper-aggressive male flies (Garbaczewska et al., 2013; Hopkins et al., 2019).

Experimental arena

Experimental arenas were described in details in Trannoy et al., 2015 (Trannoy et al., 2015a). Briefly, resin blocks containing three circular behavioral arenas (dimension: 2.3 cm diameter, 1.7 cm height) were used to conduct behavioral assays. Each arena was divided in two equal sizes by a plastic divider allowing flies to acclimate without physical interactions with the second fly. Then, dividers are removed from the arenas to start the behavioral assays. All behavioral assays were performed in empty arenas, expect for aggression assays and courtship assays with decapitated females, in which a food cup (1.5 cm diameter, 1 cm height) containing fresh standard fly food with a drop of fresh yeast paste on the surface was placed in the center of each arena.

Method details

Commonalities between all behavioral assays

Late-stage pupae were collected from stocks raised at 25°C. Male pupae were individually placed in 5ml vials (Dutscher, ref 390597) containing about 1ml of standard fly food while female pupae were placed in group of 15 in ≈47ml vials (Dutscher, ref 789001B) containing about 15ml of standard fly food. All vials were placed at 25°C for 7 days. Seven-day-old individuals were randomly inserted in arenas by negative geotaxis, and allow to acclimate for 5 min before experiments. All behavioral assays were performed at 25°C. All tests were recorded with Basler camera Ace1970 through the StreamPix8 multi video recording software (norpix.com) and analyzed with BORIS software (Friard and Gamba, 2016). After 15 min or 30 min, if pairs of flies had not yet lunged or mated respectively, we considered that no aggression or mating occurred and arenas were not considered for analysis. No power analyses were performed prior to the experiments, as we chose to use sample sizes of approximately 20 replicates, which is considered standard practice in the field.

Aggression assays

Two days prior to the aggression assays, males were anesthetized with CO2, and half were marked with a small white dot on the thorax before being returned to their original isolation vials. Aggression tests were conducted between ZT0 and ZT3 using size-matched male pairs. Males competed for access to a central food cup containing yeast. The following behaviors were recorded for 10 minutes following the first lunge: latency to lunge (i.e., the time between the initial encounter on the food cup and the first lunge), total number of lunges, and number of boxing events.

Mating assays

Seven-day-old individuals were tested in arenas without any food source. Latency to court and to mate, defined as the time between the first encounter and either the first Unilateral Wing Extension (UWE) or the initiation of mating, respectively, were measured. Total UWE duration was recorded over a 10-minute period starting from the first encounter. Mating proportion (proportion of fly pairs that succeeded to mate) and mating duration were also scored. In all assays, males were reared in social isolation, except for Figure 1H, where they were housed in groups of 10 males for 7 days.

Mating assays with decapitated females

Females were anesthetized with CO2 and placed under the binocular for head removal. Experiments with decapitated females were performed in arenas with food cup filled with fresh fly food but without yeast. The females’ bodies were gently pressed into the food cup to ensure they remained completely immobile, as males can move their bodies during courtship. This allows to maintain a constant distance between the two decapitated females across trials, thereby improving the reproducibility of the assay. Female bodies were positioned in the food cup with the abdomen raised, allowing males to initiate and complete copulation.

Multiple consecutive mating events assays

Single naïve Cs and Bully males were sequentially paired with six virgin Cs females. For each male, number of mating and each mating duration were recorded. After each mating, the mated female was transferred to a 5 ml vial containing fly food to assess progeny production.

Mating assays throughout the lifespan

24 Cs and Bully males were reared in isolation and paired with seven-day-old virgin Cs females for mating assays, conducted every seven days until all males had died. At each time point, mating duration and mating success (proportion of mating) were recorded. After all Cs males had died, females that had mated with Bully males were transferred to 5 ml vials containing fly food to assess progeny production.

Attractiveness assays

Measure of post-mating UWE

Mating assays were performed by pairing naïve Cs and Bully males with a Cs female. Right after mating end, total time that males displayed UWE for 10 minutes towards the just-mated females was scored.

Measure of post-mating UWE with switched females

Mating assays were performed with naive Cs or Bully males and virgin Cs females. Immediately after mating ends, females were switched: Cs males interacted with Bully-mated females and Bully males interacted with Cs-mated females. Then total time of UWE towards the mated females was scored during 10 min.

Measure of post-mating UWE with decapitated females

Mating assays were conducted using naïve Cs or Bully males paired with virgin Cs females. Following copulation, mated females were decapitated. One decapitated Cs-mated and one Bully-mated female were then placed on the surface of the food cup. A new naïve Cs or Bully male was introduced and allowed to interact with both decapitated females. The duration of Unilateral Wing Extensions (UWE) directed toward each female was recorded over a 2-minute period following the first encounter. Arenas in which males interacted with only one of the two females were excluded from the analysis, as no choice could be determined.

Males’ perception assays

A decapitated Cs male and female were placed on a food cup in presence of either a naïve Cs or a Bully male. The time spent by the focal male courting each decapitated fly was quantified over a 2-minute period following the first unilateral wing extension (UWE). Arenas in which the male interacted with only one of the two decapitated flies were excluded from the analysis, as no choice could be determined.

Re-mating assays

Mating assays were performed with naïve Cs or Bully males and virgin Cs females. Mated-females were then individually placed in 5ml vials containing about 1ml of standard fly food and kept at 25°C. 12 or 14 days after the initial mating, mated-females were paired with a new naïve Cs male in empty arena and the proportion of re-mating during 30 min was scored.

Survival assays

Survival with isolated males

Cs, homozygous and heterozygous Bully late-stage pupae were individually isolated in 5 ml vials containing 1ml of standard fly food and kept at 25°C until adult emergence. The number of dead individuals was recorded daily until all flies had died. To maintain food quality, flies were transferred weekly to fresh vials containing approximately 1 ml of new fly food.

Survival with group-raised males

Late-stage Cs and homozygous Bully male pupae were placed in group with Cs female pupae in ≈47ml vials containing 15mL standard fly food and kept at 25°C until adult emergence. Groups were composed of 10 males (either Cs or Bully) and 12 Cs females. The number of dead males was recorded daily until they all had died. To maintain food quality, flies were transferred twice a week to fresh vials containing approximately 15mL of new fly food.

CHCs analysis

Extraction and analysis of cuticular hydrocarbons (CHCs) of naive Cs and Bully males, as well as virgin Cs, Cs- and Bully-mated females were performed in Jean-Christophe Billeter’s lab, Groningen, Netherlands, with a Gas Chromatography coupled with Flame Ionization Detection (GC-FID) as already described (Laturney and Billeter, 2016). Briefly, single whole flies were put in a 2ml glass vial (Supelco certified vial kit, 2ml clear glass vial 12×32mm and Supelco inserts) containing 50µl of nC18+nC26 10 ng/µl standard solution (made in Billeter’s lab) diluted in n-Hexane 99%+ (Acrōs Organic®). Vials were then vortexed at the lowest speed for 2min, afterward flies’ body were gently removed from vials using a “J-shape” paper clip sliced at the end, taking care not to damage the body to prevent from hemolymph contamination. Finally, samples were placed in GC-FID for apolar compounds analysis only.

cVA quantification into genital tract

Mating assays were performed with naïve Cs or Bully males and virgin Cs female. Right after mating, females were transferred into a 5ml vial previously put in ice in −80°C fridge to make them fall asleep quickly and prevent from sperm ejection. Mated-females’ genital tract was dissected in PBS with precision forceps under a binocular loupe. Protocol was already employed in (Verschut et al., 2022). Briefly, abdomen was torn to make ovaries and genital tract out and visible, then ovaries, oviduct, spermathecae, parovaria and cuticula were removed to keep only the uterus, taking care not to injure it to prevent from sperm leak. Each uterus was inserted in a 2ml glass vial (Supelco certified vial kit, 2ml clear glass vial 12×32mm and Supelco inserts) containing 50µl of n-Hexane 99%+ (Acrōs Organic) for 1h at ambient temperature. Hexane solution was then evaporated under helium flow to prevent from cVA oxidation with dioxygen. Samples were sent to Jean-Christophe Billeter’s lab for GC-FID analysis of cVA, as describe above in CHCs analysis method part.

Quantification and statistical analysis

No data have been removed from the statistical analysis (except arenas where no choice was observed within the males’ perception assays, as described below). Some experiments were independently replicated by different authors to reduce potential bias. All statistical analyses were performed on R software version 4.0. (RC Team, 2020) with packages (Fox and Weisberg, 2019), coin (Hothorn et al., 2006), dplyr (Wickham et al., 2023), factoextra (Kassambara and Mundt, 2020), FactoMineR (Lê et al., 2008), lme4 (Bates et al., 2015), lmerTest (Kuznetsova et al., 2017), MASS (Venables and Ripley, 2002), MuMIn (Bartoń, 2023), nlme (Pinheiro et al., 2022), performance (Lüdecke et al., 2021), pgirmess (Giraudoux. P, 2018), predictmeans (Luo et al., 2021), survival (Therneau, 2022), vegan (Oksanen J, 2022).

Comparisons of means (k =2) were performed with a T-test when possible (i.e. mating duration) otherwise a Fisher Pitman permutation test was used (i.e. UWE, latency to court and to mate). For comparisons with more than 2 conditions, a linear model (LM) was used and tested either with a type II ANOVA when possible (i.e. mating duration) or a permutation test instead (i.e. latency to lunge and CHC concentrations). For comparison of proportions and counting events, a type II analysis of deviance table with a Likelihood Ratio Test (i.e. mating success, proportion of female re-mating, number of lunge and boxing events) or a Wald Chi-square test (i.e. number of consecutive mating events) was performed. A Principal Component Analysis (PCA) was performed to see different CHC profiles and tested with a PerMANOVA and a SIMPER analysis. Survival data were analysed with a Weibull regression model and a Likelihood Ratio Test.

When paired data, dependency was considered thanks to linear mixed model (LMM) with the source of dependency as random factor. Data following a distribution other than a normal distribution (e.g. binomial or Poisson distribution) were computed in a generalized linear model (GLM) (i.e. mating success, number of consecutive mating events, number of mating giving progeny, proportion of females re-mating, number of lunges and boxing events). When necessary, interactions were considered within models and tested.

All post-hoc comparisons were performed with a sequential Bonferroni correction on the alpha level using either Holm method (Holm, 1979) when few numbers of comparisons or Benjamini and Hochberg method (Benjamini and Hochberg, 1995) when more comparisons. All statistical details are given in Figure 1-table supplement 1.